---
title: "Module 2 Review: Full Forecasting Pipeline"
description: "End-of-module exercise covering the complete Module 2 toolkit — baseline, ETS, ARIMA/SARIMA, and mixed decomposition models — applied to Mexico's unemployment rate."
date: 2025-06-17
format: html
params:
  fred-data: true
categories:
  - topic-exercise
  - module-review
  - module-2
---

```{r}
#| label: setup

library(tidyverse)
library(fpp3)
library(tidyquant)
library(plotly)
```

```{r}
#| label: narsil-setup
#| include: false
source(here::here("R/narsil_theme.R"))
theme_set(theme_narsil())
```

## Introduction

This exercise brings together everything covered in Module 2. You will apply the full forecasting pipeline — from EDA through model comparison — to a new series: Mexico's quarterly unemployment rate. Your goal is to build all model families covered in Module 2, evaluate them against the Module 1 baseline, and identify which approach performs best on this series.

## The Series: Mexico Unemployment Rate

Mexico's harmonized unemployment rate (FRED ticker `LRHUTTTTMXQ156N`) is a quarterly series published by the OECD. It measures the percentage of the labor force that is unemployed and actively seeking work.

```{r}
#| label: data-load
#| message: false

unemp <- tq_get(
  "LRHUTTTTMXQ156N",
  get = "economic.data",
  from = "2005-01-01"
) |>
  mutate(date = yearquarter(date)) |>
  as_tsibble(index = date) |>
  rename(unemp = price)

unemp
```

## Exploratory Data Analysis

Before fitting any model, take time to understand the series. The plots below cover the four standard views.

```{r}
#| label: autoplot-student
#| eval: false

unemp |>
  autoplot(unemp) +
  labs(
    title = "Mexico unemployment rate",
    y     = "Unemployment rate (%)",
    x     = ""
  )
```

```{r}
#| label: autoplot-p
#| echo: false
#| output: false

unemp_p <- unemp |>
  autoplot(unemp) +
  labs(
    title = "Mexico unemployment rate",
    y     = "Unemployment rate (%)",
    x     = ""
  )
```

```{r}
#| label: autoplot-render
#| renderings: [light, dark]
#| echo: false

ggplotly(unemp_p + scale_narsil(), dynamicTicks = TRUE) |> gg_to_plotly_narsil()
ggplotly(unemp_p + scale_narsil_dark(), dynamicTicks = TRUE) |> gg_to_plotly_narsil(dark = TRUE)
```

Note the sharp spike around 2020 Q2 — the COVID-19 shock. Keep this in mind when interpreting model diagnostics and accuracy.

```{r}
#| label: season-student
#| eval: false

unemp |> gg_season(unemp)
```

```{r}
#| label: season-render
#| renderings: [light, dark]
#| echo: false
#| results: hide

theme_narsil()
unemp |> gg_season(unemp)
theme_narsil_dark()
unemp |> gg_season(unemp)
```

```{r}
#| label: subseries-student
#| eval: false

unemp |> gg_subseries(unemp)
```

```{r}
#| label: subseries-render
#| renderings: [light, dark]
#| echo: false

unemp |> gg_subseries(unemp) + scale_narsil()
unemp |> gg_subseries(unemp) + scale_narsil_dark()
```

```{r}
#| label: tsdisplay-student
#| eval: false

unemp |> gg_tsdisplay(unemp, plot_type = "partial")
```

```{r}
#| label: tsdisplay-render
#| echo: false
#| results: hide

theme_narsil()
unemp |> gg_tsdisplay(unemp, plot_type = "partial")
```

::: {.callout-note appearance="simple"}
## What to look for

- **Trend**: Is there a long-run level shift, or does the series return to a stable mean after shocks?
- **Seasonality**: Is there a repeating quarterly pattern? How stable is it across years?
- **ACF/PACF**: What correlation structure would you expect in the residuals? Does the series look stationary?
:::

## Train/Test Split

```{r}
#| label: split

h <- 8 # 2 years ahead

unemp_train <- unemp |> slice_head(n = nrow(unemp) - h)
unemp_test  <- unemp |> slice_tail(n = h)
```

The test set covers the most recent 8 quarters (2 years). All model fitting is done on `unemp_train`; accuracy is evaluated against the full `unemp` tsibble.

---

# Exercise

This is a full-pipeline exercise. Work through each section in order — later sections depend on the objects you build earlier.

## Exercise 1 — Module 1 baseline

Re-estimate the Module 1 baseline model on `unemp_train`. This is your benchmark: every model you build in this exercise must be compared against it.

The baseline uses:

- STL decomposition (choose sensible `trend(window)` and `season(window)` values for a quarterly series)
- Drift on the seasonally adjusted component
- SNAIVE on the seasonal component

Save the spec as `baseline_spec` and fit it inside a `mable` called `unemp_fit`. Then:

a. Produce forecasts for `h = 8` quarters.
b. Plot the forecast (overlay, `level = NULL`).
c. Inspect residual diagnostics with `gg_tsresiduals()`. Do the residuals look like white noise? Is the COVID spike visible?
d. Compute test-set accuracy using `accuracy(unemp_fc, unemp)`.

::: {.callout-tip}
## Hint
Remember that `accuracy()` requires the **full** tsibble — not just the test set — so that MASE and RMSSE denominators are computed correctly.
:::

## Exercise 2 — ETS models

Add at least two ETS-based approaches to `unemp_fit`:

a. **STL + ETS**: Use `decomposition_model()` with an ETS model on `season_adjust`. Suppress seasonality in the ETS spec (`season("N")`). The seasonal component defaults to SNAIVE.
b. **Full ETS**: Fit `ETS(unemp)` directly, without explicit decomposition. Let R select the best error/trend/season structure automatically.

For each model:

- Add it to the same `mable` as the baseline (a single `model()` call with all specs).
- Check residual diagnostics for the STL + ETS model.
- Include both in the accuracy comparison at the end.

## Exercise 3 — ARIMA and SARIMA

Add at least two ARIMA-based approaches:

a. **STL + ARIMA**: Use `decomposition_model()` with `ARIMA(season_adjust ~ PDQ(0,0,0))`. The seasonal component defaults to SNAIVE.
b. **Automatic SARIMA**: Fit a full seasonal ARIMA directly — `ARIMA(unemp, stepwise = FALSE, approximation = FALSE)` — without explicit decomposition.

For each model:

- Add it to the same `mable`.
- Check residual diagnostics for the STL + ARIMA model.
- Include both in the accuracy comparison.

::: {.callout-note appearance="simple"}
## Identifying a manual SARIMA candidate (optional stretch)

Use `gg_tsdisplay()` on the differenced series to identify a candidate ARIMA order manually, then add it as a third model (`sarima_manual`). Compare it against the automatic selection.
:::

## Exercise 4 — Mixed decomposition models

This section connects directly to the [Modular Forecasting](../../modules/module_2/04_modular_forecasting/modular_forecasting.qmd) class document.

Add the two mixed combinations to your `mable`:

a. **STL + ETS + ARIMA**: ETS models the seasonally adjusted component (`season_adjust`); ARIMA models the seasonal component (`season_year`).
b. **STL + ARIMA + ETS**: ARIMA models the seasonally adjusted component (`season_adjust ~ PDQ(0,0,0)`); ETS models the seasonal component (`season_year ~ trend("N")`).

Recall the constraint logic for each component:

| Component | What it contains | What to suppress |
|:--|:--|:--|
| `season_adjust` | trend + cycle + remainder | seasonality |
| `season_year` | seasonal pattern only | trend |

After fitting:

- Produce a single forecast plot with all models overlaid (`level = NULL`).
- Compute a final accuracy table with all models, sorted by RMSE.
- Answer: which approach performs best on this series? Does the winner make intuitive sense given what you observed in the EDA?

## Exercise 5 — Reflection (graduate students)

Write a short paragraph (5–8 sentences) addressing the following:

a. The COVID-19 spike in 2020 Q2 is a clear outlier. How does it affect model diagnostics and accuracy metrics? Which models appear most sensitive to it?
b. The baseline was designed in Module 1 for a different series (`mexretail`). Is it still a competitive benchmark here, or does the unemployment series favor a different model family? What characteristics of the series drive your answer?
c. Would you expect the same model ranking to hold on a different country's unemployment series? What would you check first before deciding?
