---
title: "ARIMA Models"
format:
  html:
    other-links:
      - text: FPP3 Chapter 9
        href: https://otexts.com/fpp3/arima.html
  revealjs:
    output-file: arima_pres.html
---
```{r}
#| label: pkgs
#| message: false
#| include: false

library(tidyverse)
library(fpp3)
```

:::{.content-visible unless-format="revealjs"}
```{r}
#| label: pkgs_special
#| message: false

library(tidyquant) #<1>
library(plotly)    #<1>
```

1. In addition to the regular packages, here we'll use `tidyquant` and `plotly`.

:::
```{r}
#| label: data-setup
#| include: false
#| message: false

mexretail <- tq_get(
  "MEXSLRTTO01IXOBM",
  get = "economic.data",
  from = "1985-01-01"
) |>
  mutate(date = yearmonth(date)) |>
  rename(y = price) |>
  as_tsibble(index = date)

lambda <- mexretail |>
  features(y, features = guerrero) |>
  pull(lambda_guerrero)
```

# ARIMA Models

::: {.content-hidden unless-format="revealjs"}
## From AR/MA to ARIMA
:::

In the [previous topic](../02_stationarity/stationarity.qmd#ar-and-ma-models) we introduced the two basic building blocks for modeling the correlation structure of a stationary time series:

- **AR($p$):** forecast $y_t$ using its own $p$ past values.
- **MA($q$):** forecast $y_t$ using $q$ past forecast errors.

::: {.fragment}

In practice, neither model alone is usually sufficient. Real series often show patterns that require *both* autoregressive and moving average components simultaneously. And most real series are also **non-stationary** — they have trend, seasonality, or both.

:::

::: {.fragment}
::: {.callout-note appearance="simple"}
## The natural next step

If we could combine AR and MA terms *and* relax the stationarity requirement by incorporating differencing directly into the model, we would have a single, unified framework for a very wide range of time series.

That framework is **ARIMA**.
:::
:::

## The ARIMA Equation

**ARIMA** stands for **A**uto**R**egressive **I**ntegrated **M**oving **A**verage. The "Integrated" part refers to the reverse operation of differencing: if we difference a series $d$ times to make it stationary, the model is said to be integrated of order $d$.

A non-seasonal ARIMA model is written as **ARIMA($p$, $d$, $q$)**, where:

::: {.incremental}
- $p$ — order of the **autoregressive** component (how many past values)
- $d$ — order of **differencing** (how many times we difference to achieve stationarity)
- $q$ — order of the **moving average** component (how many past errors)
:::

::: {.fragment}

The model equation, written on the **differenced series** $y'_t$, is:

$$
y'_t = c + \phi_1 y'_{t-1} + \cdots + \phi_p y'_{t-p} + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q} + \varepsilon_t
$$

where $y'_t$ denotes the $d$-times differenced series and $\varepsilon_t$ is white noise.

:::

## Backshift Notation

Recall from the [previous topic](../02_stationarity/stationarity.qmd#backshift-notation) that the backshift operator $B$ satisfies $By_t = y_{t-1}$, and that $d$-th order differencing can be written as $(1-B)^d y_t$.

Using this notation, we can write the AR and MA components compactly as well:

::: {.fragment}

**AR component:**
$$\phi_1 y_{t-1} + \phi_2 y_{t-2} + \cdots + \phi_p y_{t-p} = \phi(B)\,y_t$$

where $\phi(B) = \phi_1 B + \phi_2 B^2 + \cdots + \phi_p B^p$.

**MA component:**
$$\theta_1 \varepsilon_{t-1} + \theta_2 \varepsilon_{t-2} + \cdots + \theta_q \varepsilon_{t-q} = \theta(B)\,\varepsilon_t$$

where $\theta(B) = \theta_1 B + \theta_2 B^2 + \cdots + \theta_q B^q$.

:::

::: {.fragment}

**Full ARIMA($p$, $d$, $q$) in backshift notation:**

$$
\underbrace{(1 - \phi_1 B - \cdots - \phi_p B^p)}_{\text{AR}(p)} \underbrace{(1-B)^d}_{\text{differencing}} y_t = c + \underbrace{(1 + \theta_1 B + \cdots + \theta_q B^q)}_{\text{MA}(q)} \varepsilon_t
$$

This compact form shows clearly how the three components interact. The differencing operator $(1-B)^d$ is applied first, and the AR and MA operators act on the resulting stationary series.

:::

## Special Cases

Several models we have already encountered are special cases of the ARIMA family:

::: {.fragment}

| Model | ARIMA specification |
|:------|:-------------------|
| White noise | ARIMA(0, 0, 0) without constant |
| Random walk | ARIMA(0, 1, 0) without constant |
| Random walk with drift | ARIMA(0, 1, 0) with constant |
| AR($p$) | ARIMA($p$, 0, 0) |
| MA($q$) | ARIMA(0, 0, $q$) |

:::

::: {.fragment}
::: {.callout-tip appearance="simple"}
## Effect of $c$ and $d$ on long-run forecasts

The constant $c$ and the differencing order $d$ interact to determine the long-run behavior of ARIMA forecasts:

- $d = 0$, $c = 0$: forecasts go to zero.
- $d = 0$, $c \neq 0$: forecasts go to the mean of the series.
- $d = 1$, $c = 0$: forecasts approach a non-zero constant.
- $d = 1$, $c \neq 0$: forecasts follow a straight line.
- $d = 2$, $c \neq 0$: forecasts follow a quadratic trend.

::: {.fragment}
Higher values of $d$ also produce wider prediction intervals that grow faster.
:::
:::
:::

# ARIMA in R

## Choosing the Orders

Selecting the right ARIMA order involves three decisions:

::: {.incremental}
1. **Choose $d$** — determine how many differences are needed using the stationarity protocol from the [previous topic](../02_stationarity/stationarity.qmd#unit-root-tests): Box-Cox / log transformation to stabilize variance, `unitroot_nsdiffs()` for seasonal differencing (if any), then `unitroot_ndiffs()` on the differenced series.

2. **Choose $p$ or $q$** — read the ACF and PACF of the (differenced, stationary) series:
   - **PACF cuts off** after lag $p$ → AR($p$) component.
   - **ACF cuts off** after lag $q$ → MA($q$) component.
   - Both decay gradually → mixed ARMA, try several combinations.

3. **Compare candidates** using information criteria (see [Model Selection](#sec-model-selection)).
:::

::: {.fragment}
::: {.callout-warning appearance="simple"}
## ACF/PACF reading is a starting point, not a definitive answer

Real data rarely produces textbook-perfect ACF/PACF shapes. Use the plots to narrow down candidate models, then compare them formally with AIC$_c$.
:::
:::

## Manual and Automatic Specification {#sec-manual}

In `fable`, ARIMA models are specified inside `model()` using the `ARIMA()` function. The order is controlled with `pdq()` for non-seasonal terms and `PDQ()` for seasonal terms. When `pdq()` is left unspecified, `fable` searches for the best order automatically.
```{r}
#| label: us-change-plot
#| echo: false
#| message: false

p <- us_change |>
  autoplot(Consumption, color = "steelblue") +
  labs(
    title = "US consumption — quarterly % change",
    y = "% change",
    x = ""
  )

ggplotly(p, dynamicTicks = TRUE)
```

::: {.fragment}
```{r}
#| label: us-change-tsdisplay
#| warning: false
#| message: false
#| echo: false

us_change |>
  gg_tsdisplay(Consumption, plot_type = "partial", lag_max = 16)
```

The series appears stationary ($d = 0$). The PACF shows significant spikes at lags 1–3, suggesting AR(3). We can fit this manually, compare it against automatic selection strategies, and let AIC$_c$ arbitrate — all within a single `mable`:

:::

::: {.fragment}
```{r}
#| label: us-change-fit-all
#| message: false

us_change_fit <- us_change |>
  model(
    manual     = ARIMA(Consumption ~ pdq(3, 0, 0) + PDQ(0, 0, 0)),     #<1>
    auto       = ARIMA(Consumption ~ PDQ(0, 0, 0)),                     #<2>
    semi_auto  = ARIMA(Consumption ~ pdq(1:3, 0, 0:2) + PDQ(0, 0, 0)), #<3>
    exhaustive = ARIMA(Consumption ~ PDQ(0, 0, 0),
                       stepwise      = FALSE,                           #<4>
                       approximation = FALSE)                           #<5>
  )

glance(us_change_fit) |>
  arrange(AICc) |>
  select(.model, AIC, AICc, BIC)
```

1. Manual specification using `pdq()`. `PDQ(0,0,0)` explicitly forces a non-seasonal model.
2. Leaving `pdq()` unspecified triggers automatic stepwise order selection.
3. Evaluates all combinations of $p \in \{1,2,3\}$ and $q \in \{0,1,2\}$, returns the lowest AIC$_c$.
4. `stepwise = FALSE` explores all combinations rather than navigating step by step.
5. `approximation = FALSE` uses exact likelihood. More accurate but slower on long series.

:::

::: {.fragment}
::: {.callout-note appearance="simple"}
## How does automatic selection work?

`fable` implements a stepwise search algorithm (Hyndman & Khandakar, 2008) that:

1. Determines $d$ using unit root tests.
2. Starts from a simple model and navigates through the order space by comparing AIC$_c$ values.
3. Stops when no neighboring model improves AIC$_c$.

::: {.fragment}
**Important:** this is a heuristic search — it does *not* evaluate all possible combinations. It is fast and usually good, but not guaranteed to find the global optimum.
:::
:::
:::

::: {.fragment}
::: {.callout-tip appearance="simple"}
For exploratory work, automatic selection is fine. For a final model, use `stepwise = FALSE, approximation = FALSE` and compare the result with your manually identified candidate.
:::
:::

## Model Selection {#sec-model-selection}

:::{.content-visible unless-format="revealjs"}

When we fit a statistical model, we face a fundamental tension: a more complex model (more parameters) will always fit the observed data better, but it may simply be memorizing noise rather than capturing genuine structure. This is the **bias-variance tradeoff**.

**Information criteria** address this by penalizing model complexity — they reward goodness of fit but add a cost for each additional parameter estimated. The goal is to find the model that generalizes best, not the one that fits best in-sample.

:::

When comparing candidate ARIMA models, we use three related criteria:

- **AIC** — Akaike Information Criterion
- **AIC$_c$** — AIC corrected for small sample sizes
- **BIC** — Bayesian Information Criterion (also known as Schwarz criterion)

$$
\text{AIC} = -2\log(L) + 2(p + q + k + 1)
$$

$$
\text{AIC}_c = \text{AIC} + \frac{2(p + q + k + 1)(p + q + k + 2)}{T - p - q - k - 2}
$$

$$
\text{BIC} = \text{AIC} + [\log(T) - 2](p + q + k + 1)
$$

where $k = 1$ if the model includes a constant, $k = 0$ otherwise. **Lower is better** for all three. The preferred criterion is AIC$_c$ — it converges to AIC as $T$ grows, but is more conservative with small samples.

:::{.content-visible unless-format="revealjs"}

::: {.callout-tip collapse="true"}
## AIC vs. BIC — which to prefer?

Both penalize complexity, but BIC applies a heavier penalty (it grows with $\log T$ rather than a fixed 2). In practice:

- **AIC$_c$** tends to select slightly richer models and is generally preferred for forecasting.
- **BIC** tends to select more parsimonious models and is preferred when interpretability matters.

For this course, use **AIC$_c$** as the default.
:::

:::

::: {.fragment}
::: {.callout-important appearance="simple"}
## Information criteria require the same scale

Information criteria are only comparable across models fitted to **identical data** — same transformation, same differencing order. You cannot compare AIC$_c$ of an ARIMA fitted to $\log(y_t)$ vs. one fitted to $y_t$, or ARIMA(1,1,1) vs. ARIMA(1,0,1). When models differ in $d$, use forecast accuracy on a test set instead.
:::
:::

## Box-Jenkins Methodology

ARIMA model building follows an iterative procedure known as the **Box-Jenkins methodology**:

::: {.incremental}
1. **Plot the data.** Identify unusual observations, structural breaks, outliers.
2. **Stabilize the variance** if needed — apply a log or Box-Cox transformation.
3. **Determine $d$** — use `unitroot_nsdiffs()` and `unitroot_ndiffs()`.
4. **Examine ACF and PACF** of the stationary series to identify candidate orders $p$ and $q$.
5. **Fit candidate models** and compare using AIC$_c$.
6. **Check residuals** — ACF of residuals and Ljung-Box test. If not white noise, refine.
7. **Forecast** once residuals pass diagnostic checks.
:::

::: {.fragment}
::: {.callout-note appearance="simple"}
## It's iterative

The Box-Jenkins procedure is a cycle, not a checklist. It is normal to go through steps 4–6 two or three times before settling on a model.
:::
:::

## Residual Diagnostics

We have already used the Ljung-Box test to check residuals from benchmark models. When applying it to **ARIMA models**, there is one important adjustment: you must account for the degrees of freedom consumed by the estimated AR and MA parameters.
```{r}
#| label: us-change-residuals
#| warning: false

us_change_fit |>
  select(manual) |>
  gg_tsresiduals()
```
```{r}
#| label: us-change-ljungbox

us_change_fit |>
  select(manual) |>
  augment() |>
  features(.innov, ljung_box,
           lag = 10,
           dof = us_change_fit |>                          #<1>
                   select(manual) |>
                   coefficients() |>
                   filter(str_detect(term, "^ar|^ma")) |>
                   nrow())
```

1. `dof` is computed dynamically as the number of estimated AR and MA coefficients, so it always matches the fitted model regardless of order.

## Forecasting

Once the model passes residual diagnostics, forecasting uses the same `forecast()` workflow as all other `fable` models:
```{r}
#| label: us-change-forecast
#| message: false
#| warning: false

us_change_fit |>
  select(manual) |>
  forecast(h = 10) |>
  autoplot(us_change |> filter(year(Quarter) >= 2005)) +
  labs(
    title = "US consumption — ARIMA(3,0,0) forecast",
    y = "% change"
  )
```

::: {.fragment}
::: {.callout-note appearance="simple"}
## Prediction intervals

ARIMA prediction intervals are derived analytically from the model's MA representation. Their width grows with the forecast horizon, and **grows faster as $d$ increases** — a direct consequence of accumulated uncertainty from differencing. For a random walk ($d = 1$), intervals widen at rate $\sqrt{h}$; for $d = 2$, they widen even faster.
:::
:::

# Seasonal ARIMA

## STL + ARIMA

`mexretail` is a monthly series with strong trend and seasonality. We already know how to handle it with STL decomposition. In Module 1 we combined STL with SNAIVE; in Module 2 we replaced SNAIVE with ETS. The natural next step is to replace it with ARIMA.

The idea is the same: use STL to handle the seasonal pattern, and let ARIMA model the trend-cycle on the **seasonally adjusted series**.
```{r}
#| label: stl-arima-fit
#| message: false

mexretail_fit_stl_arima <- mexretail |>
  model(
    stl_arima = decomposition_model(       #<1>
      STL(box_cox(y, lambda) ~             #<2>
            trend(window = NULL) +
            season(window = "periodic"),
          robust = TRUE),
      ARIMA(season_adjust + PDQ(0,0,0))    #<3>
    )
  )

report(mexretail_fit_stl_arima)
```

1. `decomposition_model()` combines a decomposition method with a model for one or more of its components.
2. STL decomposes the Box-Cox-transformed series. The seasonal component is handled implicitly by the decomposition.
3. `ARIMA(season_adjust)` fits an ARIMA model automatically to the seasonally adjusted component. The seasonal part is re-added when generating forecasts.
```{r}
#| label: stl-arima-residuals
#| warning: false

mexretail_fit_stl_arima |> gg_tsresiduals()
```
```{r}
#| label: stl-arima-forecast
#| message: false
#| warning: false

mexretail_fit_stl_arima |>
  forecast(h = 24) |>
  autoplot(mexretail |> filter(year(date) >= 2015)) +
  labs(
    title = "mexretail — STL + ARIMA forecast",
    y = "Index 2015=100"
  )
```

## SARIMA

STL + ARIMA is a two-stage approach: decompose first, then model. An alternative is to handle seasonality **within a single ARIMA model** by adding seasonal AR and MA terms. This is a **Seasonal ARIMA** model, written as:

$$
\text{ARIMA}\underbrace{(p,\,d,\,q)}_{\text{non-seasonal}}\underbrace{(P,\,D,\,Q)_{m}}_{\text{seasonal}}
$$

where $m$ is the seasonal period ($m = 12$ for monthly data).

::: {.fragment}

The full model in backshift notation is:

$$
\underbrace{\Phi(B^m)}_{\text{seasonal AR}} \underbrace{\phi(B)}_{\text{non-seasonal AR}} \underbrace{(1-B^m)^D (1-B)^d}_{\text{differencing}} y_t = c + \underbrace{\Theta(B^m)}_{\text{seasonal MA}} \underbrace{\theta(B)}_{\text{non-seasonal MA}} \varepsilon_t
$$

:::

::: {.content-hidden unless-format="revealjs"}
## Reading ACF/PACF for Seasonal Orders
:::

::: {.content-visible unless-format="revealjs"}
### Reading ACF/PACF for Seasonal Orders
:::

The non-seasonal orders ($p$, $q$) are read from the first few lags of the ACF/PACF, as before. The seasonal orders ($P$, $Q$) are read from the **seasonal lags** ($m$, $2m$, $3m$, ...):

| Pattern at seasonal lags | Suggested component |
|:--------------------------|:--------------------|
| PACF spikes cut off after lag $Pm$ | Seasonal AR($P$) |
| ACF spikes cut off after lag $Qm$ | Seasonal MA($Q$) |
| Both decay gradually | Mixed seasonal ARMA |

::: {.fragment}
::: {.callout-tip appearance="simple"}
Focus on the seasonal lags *separately* from the non-seasonal lags. Use `lag_max = 3m` or `4m` to make the seasonal structure clearly visible.
:::
:::

::: {.content-hidden unless-format="revealjs"}
## SARIMA for `mexretail`
:::

::: {.content-visible unless-format="revealjs"}
### SARIMA for `mexretail`
:::

The stationarity protocol for `mexretail` established that we need $D = 1$ and $d = 1$ applied to the Box-Cox-transformed series. Let's inspect the ACF/PACF to choose $p$, $q$, $P$, $Q$:
```{r}
#| label: mexretail-tsdisplay
#| warning: false
#| message: false

mexretail |>
  gg_tsdisplay(
    difference(box_cox(y, lambda), 12) |> difference(1),
    plot_type = "partial",
    lag_max   = 48
  )
```

::: {.fragment}
::: {.callout-note appearance="simple"}
## Reading the plot

- **Non-seasonal lags (1–11):**
  - ACF: spike at lag 1, and up to lag 4 → try $q = 1$ up to $q = 4$.
  - PACF: spikes at lags 1 and 2, borderline at lag 3 → try $p = 2$ or $p = 3$.
- **Seasonal lags (12, 24, 36, 48):**
  - ACF: spike at lag 12, barely at lag 24 → try $Q = 1$ or $Q = 2$.
  - PACF: spikes at lags 12, 24, and slightly at 36 and 48 → try $P = 2$, $P = 3$, or even $P = 4$.
:::
:::

::: {.fragment}
::: {.callout-tip appearance="simple"}
## When ACF/PACF suggest large orders, start simple

A strict reading of the ACF/PACF above could justify orders as large as AR(3), MA(4), SAR(4), SMA(2) — a model with over a dozen parameters. In practice, this is rarely a good starting point: high-order models are harder to estimate, more prone to overfitting, and — if specified manually — can fail to converge or return non-invertible roots altogether.

A better strategy: start with a parsimonious candidate like $\text{ARIMA}(0,1,1)(0,1,1)_{12}$ or $\text{ARIMA}(1,1,1)(1,1,1)_{12}$, fit a few of the larger candidates suggested by the plots, and let AIC$_c$ arbitrate. You will often find that the simpler model performs comparably or better — and it will always be more stable.
:::
:::
```{r}
#| label: sarima-fit
#| message: false

mexretail_fit_sarima <- mexretail |>
  model(
    sarima_011_110 = ARIMA(box_cox(y, lambda) ~
                             pdq(0, 1, 1) + PDQ(1, 1, 0)), #<1>
    sarima_011_011 = ARIMA(box_cox(y, lambda) ~
                             pdq(0, 1, 1) + PDQ(0, 1, 1)), #<2>
    sarima_auto    = ARIMA(box_cox(y, lambda),              #<3>
                           stepwise      = FALSE,
                           approximation = FALSE)
  )

glance(mexretail_fit_sarima) |>
  arrange(AICc) |>
  select(.model, AIC, AICc, BIC)
```

1. First candidate from ACF/PACF: non-seasonal MA(1), seasonal AR(1).
2. Second candidate: non-seasonal MA(1), seasonal MA(1) — a very common structure for monthly economic data.
3. Exhaustive automatic search for comparison.
```{r}
#| label: sarima-report
#| message: false

mexretail_fit_sarima |>
  select(sarima_auto) |>
  report()
```
```{r}
#| label: sarima-residuals
#| warning: false

mexretail_fit_sarima |>
  select(sarima_auto) |>
  gg_tsresiduals(lag_max = 48)
```
```{r}
#| label: sarima-ljungbox

mexretail_fit_sarima |>
  select(sarima_auto) |>
  augment() |>
  features(.innov, ljung_box,
           lag = 24,
           dof = mexretail_fit_sarima |>                        #<1>
                   select(sarima_auto) |>
                   coefficients() |>
                   filter(str_detect(term, "^ar|^ma|^sar|^sma")) |>
                   nrow())
```

1. `dof` is computed dynamically as $p + q + P + Q$ for the fitted model, so it always reflects the actual order selected by the exhaustive search.

# STL + ARIMA vs. SARIMA

::: {.content-hidden unless-format="revealjs"}
## Comparing Both Strategies
:::

Both approaches are valid for seasonal series. Let's compare them on `mexretail` using a train/test split:
```{r}
#| label: train-test-split
#| message: false

mexretail_train <- mexretail |> filter(year(date) <= 2019)
mexretail_test  <- mexretail |> filter(year(date) >  2019)
```
```{r}
#| label: comparison-fit
#| message: false
#| warning: false

mexretail_fit <- mexretail_train |>
  model(
    stl_arima = decomposition_model(
      STL(box_cox(y, lambda) ~
            trend(window = NULL) +
            season(window = "periodic"),
          robust = TRUE),
      ARIMA(season_adjust)
    ),
    sarima = ARIMA(box_cox(y, lambda),
                   stepwise      = FALSE,
                   approximation = FALSE)
  )
```
```{r}
#| label: comparison-accuracy
#| message: false
#| warning: false

mexretail_fit |>
  forecast(h = nrow(mexretail_test)) |>
  accuracy(mexretail) |>
  select(.model, RMSE, MAE, MAPE, MASE) |>
  arrange(RMSE)
```
```{r}
#| label: comparison-plot-facet
#| message: false
#| warning: false
#| echo: false

p <- mexretail_fit |>
  forecast(h = nrow(mexretail_test)) |>
  autoplot(mexretail |> filter(year(date) >= 2015), level = 80) +
  labs(
    title    = "mexretail — STL + ARIMA vs. SARIMA",
    y        = "Index 2015=100",
    subtitle = "80% prediction intervals"
  )

p +
  facet_wrap(vars(.model), ncol = 1)
```
```{r}
#| label: comparison-plot-overlay
#| message: false
#| warning: false
#| echo: false

p
```

::: {.fragment}
::: {.callout-note appearance="simple"}
## When to use which

| | STL + ARIMA | SARIMA |
|:--|:--|:--|
| **Seasonal pattern** | Flexible — can change over time | Fixed structure |
| **Outlier robustness** | High (with `robust = TRUE`) | Lower |
| **Interpretability** | Decomposition is visible | Single compact model |
| **Multiple seasonality** | Handles it naturally | Difficult |
| **Parsimony** | Two models in sequence | One unified model |

Neither approach dominates universally. Let the data and the accuracy metrics guide the choice.
:::
:::