---
title: "Dynamic & Harmonic Regression"
format:
  html:
    other-links:
      - text: FPP3 Chapter 10
        href: https://otexts.com/fpp3/dynamic.html
  revealjs:
    output-file: dynamic_regression_pres.html
freeze: auto
---

:::{.content-visible unless-format="revealjs"}

```{r}
#| label: pkgs
#| message: false
#| include: false

library(tidyverse)
library(fpp3)
```

```{r}
#| label: pkgs_special
#| message: false

library(plotly) #<1>
```

1. In addition to the regular packages, here we'll use `plotly` for interactive plots.

:::

```{r}
#| label: data-setup
#| include: false
#| message: false

us_change <- fpp3::us_change

us_change_fit_mult <- us_change |>
  model(
    multiple = TSLM(Consumption ~ Income + Production + Savings + Unemployment)
  )

vic_elec_daily <- vic_elec |>
  filter(year(Time) == 2014) |>
  index_by(Date = date(Time)) |>
  summarise(
    Demand      = sum(Demand) / 1e3,
    Temperature = max(Temperature),
    Holiday     = any(Holiday)
  ) |>
  mutate(
    Day_Type = case_when(
      Holiday              ~ "Holiday",
      wday(Date) %in% 2:6  ~ "Weekday",
      TRUE                 ~ "Weekend"
    )
  )
```

# One Assumption Left to Fix

:::{.content-hidden unless-format="revealjs"}

## Where we left off

:::

In the [previous session](../02_regression/regression.qmd), we built a multiple regression model for US consumption:

$$y_{t,\text{Cons.}} = \beta_0 + \beta_1 x_{t,\text{Inc.}} + \beta_2 x_{t,\text{Prod.}} + \beta_3 x_{t,\text{Sav.}} + \beta_4 x_{t,\text{Unemp.}} + \varepsilon_t$$

The model improved substantially over the simple version — $\bar{R}^2$ went from ~0.15 to ~0.75, and the residuals looked much closer to white noise.

:::{.fragment}

But "much closer" is not the same as "white noise." Let's check:

```{r}
#| label: mult-resid-check

us_change_fit_mult |>
  augment() |>
  features(.resid, ljung_box, lag = 10, dof = 5)
```

:::

:::{.fragment}

The p-value is not zero, but in practice with real data there is nearly always *some* residual autocorrelation left after TSLM. And that matters:

:::{.callout-warning appearance="simple"}
## What autocorrelated residuals cost us

If $\varepsilon_t$ is autocorrelated, OLS is still unbiased — but standard errors are wrong, confidence intervals are wrong, and **prediction intervals are wrong**. The model is leaving structure on the table that could be captured.
:::

:::

:::{.content-hidden unless-format="revealjs"}

##

:::

The culprit is the **white noise assumption** on $\varepsilon_t$:

$$\varepsilon_t \overset{\text{iid}}{\sim} N(0, \sigma^2)$$

What if instead of forcing this, we *modeled* whatever autocorrelation remains?

:::{.fragment}

:::{.callout-note appearance="simple"}
## The upgrade: relax the white noise assumption

| Model | Error term |
|:---|:---|
| TSLM | $\varepsilon_t \overset{\text{iid}}{\sim} N(0,\sigma^2)$ — must be white noise |
| **Dynamic regression** | $\eta_t \sim \text{ARIMA}(p,d,q)$ — **can be autocorrelated** |

Same predictors. Same regression structure. One assumption relaxed.
:::

:::

# Regression with ARIMA Errors

:::{.content-hidden unless-format="revealjs"}

## The Model

:::

The dynamic regression model replaces the white noise error $\varepsilon_t$ with $\eta_t$, which is allowed to follow an ARIMA process:

$$y_t = \beta_0 + \beta_1 x_{1,t} + \cdots + \beta_k x_{k,t} + \eta_t$$

$$\eta_t \sim \text{ARIMA}(p, d, q)(P, D, Q)_m$$

:::{.fragment}

The model still has white noise — it just lives one level deeper. There are now **two distinct error terms**:

:::{.callout-note appearance="simple"}
## Two error terms

- $\eta_t$ — the **regression error**: how far $y_t$ is from the regression surface. Still autocorrelated, now modeled explicitly.
- $\varepsilon_t$ — the **ARIMA innovation**: the white noise that drives $\eta_t$. This is the only truly random component.

We diagnose the model using $\hat{\varepsilon}_t$, not $\hat{\eta}_t$.
:::

:::

:::{.content-hidden unless-format="revealjs"}

## Stationarity

:::

:::{.content-visible unless-format="revealjs"}

### Stationarity

:::

:::{.callout-warning appearance="simple"}
## All variables must be stationary

Dynamic regression requires that both $y_t$ and all predictors $x_{k,t}$ are stationary. If any variable needs differencing, apply it to **all variables simultaneously**. Specifying `pdq(d = 1)` does this automatically.

For `us_change`, all variables are already percentage changes — stationary by construction.
:::

:::{.content-hidden unless-format="revealjs"}

## Fitting in R

:::

:::{.content-visible unless-format="revealjs"}

### Fitting in R

:::

`ARIMA()` handles dynamic regression by simply adding predictors to the right-hand side of the formula — the same syntax as `TSLM()`:

```{r}
#| eval: false

# TSLM — white noise errors assumed
TSLM(y ~ x1 + x2)

# Dynamic regression — ARIMA errors, order chosen automatically
ARIMA(y ~ x1 + x2)

# Force differencing on all variables
ARIMA(y ~ x1 + x2 + pdq(d = 1))
```

# Example: US Consumption

:::{.content-hidden unless-format="revealjs"}

## TSLM → Dynamic Regression {auto-animate="true"}

```{r}
#| label: us-tslm-show
#| eval: false

us_change |>
  model(
    tslm    = TSLM(Consumption ~ Income + Production + Savings + Unemployment)
  )
```

## TSLM → Dynamic Regression {auto-animate="true"}

```{r}
#| label: us-both-show
#| eval: false

us_change |>
  model(
    tslm    = TSLM(Consumption ~ Income + Production + Savings + Unemployment),
    dynamic = ARIMA(Consumption ~ Income + Production + Savings + Unemployment)
  )
```

:::

:::{.content-hidden unless-format="revealjs"}

## Fitting the Dynamic Regression {auto-animate="true"}

:::

:::{.content-visible unless-format="revealjs"}

### Fitting the Dynamic Regression

:::

```{r}
#| label: us-dynamic-fit

us_change_fit_dyn <- us_change |>
  model(
    tslm    = TSLM(Consumption ~ Income + Production +      #<1>
                     Savings + Unemployment),
    dynamic = ARIMA(Consumption ~ Income + Production +     #<2>
                      Savings + Unemployment)
  )

us_change_fit_dyn |>
  select(dynamic) |>
  report()
```

1. Our previous model — kept here for direct comparison.
2. Same predictors, same formula structure. `ARIMA()` selects the error order automatically via AIC$_c$.

:::{.content-hidden unless-format="revealjs"}

## Two Types of Residuals

:::

:::{.content-visible unless-format="revealjs"}

### Two Types of Residuals

:::

We can extract both error types from the fitted model:

```{r}
#| label: us-two-residuals

bind_rows(
  `Regression errors` = as_tibble(                               #<1>
    residuals(us_change_fit_dyn |> select(dynamic),
              type = "regression")),
  `ARIMA innovations` = as_tibble(                               #<2>
    residuals(us_change_fit_dyn |> select(dynamic),
              type = "innovation")),
  .id = "type"
) |>
  ggplot(aes(x = Quarter, y = .resid)) +
  geom_line() +
  facet_wrap(~ type, nrow = 2, scales = "free_y") +
  labs(y = NULL, x = NULL)
```

1. `type = "regression"` → $\hat{\eta}_t$: the regression residuals. These still have autocorrelation structure, now captured by the ARIMA component.
2. `type = "innovation"` → $\hat{\varepsilon}_t$: what should be white noise. This is what we diagnose.

:::{.content-hidden unless-format="revealjs"}

## Residual Diagnostics

:::

:::{.content-visible unless-format="revealjs"}

### Residual Diagnostics

:::

```{r}
#| label: us-diagnostics

us_change_fit_dyn |>
  select(dynamic) |>
  gg_tsresiduals()
```

```{r}
#| label: us-ljungbox

us_change_fit_dyn |>
  select(dynamic) |>
  augment() |>
  features(.innov, ljung_box,
           lag = 10,
           dof = us_change_fit_dyn |>                   #<1>
                   select(dynamic) |>
                   coefficients() |>
                   filter(str_detect(term, "^ar|^ma")) |>
                   nrow())
```

1. `dof` is computed dynamically as the number of estimated AR and MA coefficients — always matches the fitted model regardless of order selected.

:::{.content-hidden unless-format="revealjs"}

## Does It Actually Improve Things?

:::

:::{.content-visible unless-format="revealjs"}

### Does It Actually Improve Things?

:::

Let's compare the two models side by side on a held-out test set:

```{r}
#| label: us-accuracy-compare
#| message: false

us_change_train <- us_change |> filter_index(. ~ "2015 Q4")
us_change_test  <- us_change |> filter_index("2016 Q1" ~ .)

us_change_fit_cv <- us_change_train |>
  model(
    tslm    = TSLM(Consumption ~ Income + Production + Savings + Unemployment),
    dynamic = ARIMA(Consumption ~ Income + Production + Savings + Unemployment)
  )

us_change_fit_cv |>
  forecast(new_data = us_change_test) |>
  accuracy(us_change_test) |>
  select(.model, RMSE, MAE, MAPE)
```

:::{.callout-important appearance="simple"}
## Key insight

The dynamic regression model passes the Ljung-Box test — something the TSLM could not. By modeling the residual autocorrelation explicitly with ARIMA, we get **valid standard errors**, **correct prediction intervals**, and typically better out-of-sample accuracy.
:::

# Forecasting with Dynamic Regression

:::{.content-hidden unless-format="revealjs"}

## What changes vs. TSLM?

:::

Forecasting works exactly as in [Session 3.2](../02_regression/regression.qmd#forecasting-with-regression) — you still need future values of the predictors, and the same three strategies apply (ex-ante, ex-post, scenario-based).

:::{.fragment}

The only practical difference: you use `new_data()` to build the future dataset and pass it to `forecast()`. The ARIMA component of the errors is projected forward automatically.

:::

:::{.content-hidden unless-format="revealjs"}

## Scenario Forecast {auto-animate="true"}

```{r}
#| label: us-future-1

us_change_future <- new_data(us_change, 8)
us_change_future
```

## Scenario Forecast {auto-animate="true"}

```{r}
#| label: us-future-2

us_change_future <- new_data(us_change, 8) |>
  mutate(
    Income       = mean(us_change$Income),
    Production   = mean(us_change$Production),
    Savings      = mean(us_change$Savings),
    Unemployment = mean(us_change$Unemployment)
  )
us_change_future
```

:::

:::{.content-hidden unless-format="revealjs"}

## Scenario Forecast {auto-animate="true"}

:::

:::{.content-visible unless-format="revealjs"}

### Scenario Forecast

:::

```{r}
#| label: us-forecast-full

us_change_future <- new_data(us_change, 8) |>          #<1>
  mutate(
    Income       = mean(us_change$Income),             #<2>
    Production   = mean(us_change$Production),
    Savings      = mean(us_change$Savings),
    Unemployment = mean(us_change$Unemployment)
  )

us_change_fit_dyn |>
  select(dynamic) |>
  forecast(new_data = us_change_future) |>
  autoplot(tail(us_change, 60)) +
  labs(
    title = "US Consumption: dynamic regression forecast",
    subtitle = "Baseline scenario: all predictors at their historical mean",
    y = "% change", x = NULL
  )
```

1. `new_data()` creates 8 future rows preserving the tsibble index.
2. Baseline scenario: each predictor held at its long-run average.

:::{.content-hidden unless-format="revealjs"}

## TSLM vs. Dynamic: Prediction Intervals

:::

:::{.content-visible unless-format="revealjs"}

### TSLM vs. Dynamic: Prediction Intervals

:::

```{r}
#| label: us-interval-compare
#| message: false
#| echo: false

us_change_fit_dyn |>
  forecast(new_data = us_change_future) |>
  autoplot(tail(us_change, 60), level = 80) +
  facet_wrap(~ .model) +
  labs(
    title = "Modeling ARIMA errors narrows prediction intervals",
    subtitle = "80% prediction intervals shown",
    y = "% change", x = NULL
  )
```

:::{.callout-warning appearance="simple"}
## The prediction interval caveat

Dynamic regression intervals assume future predictor values are **known exactly**. When they are estimated (as in the scenario above), true uncertainty is larger than what the intervals show. This applies equally to TSLM — dynamic regression does not make this problem worse, it just fixes the autocorrelation problem.
:::

# A Second Example: Electricity Demand

:::{.content-hidden unless-format="revealjs"}

## Why a Second Example?

:::

The `us_change` example was clean: stationary series, linear relationship, one seasonal period. Electricity demand pushes two things further:

:::{.incremental}

- A **nonlinear** relationship between the predictor and the response
- A **categorical** predictor (day type) alongside a continuous one

:::

:::{.content-hidden unless-format="revealjs"}

## The Data

:::

:::{.content-visible unless-format="revealjs"}

### The Data

:::

Daily electricity demand in Victoria, Australia (2014):

```{r}
#| label: elec-scatter
#| echo: false

p <- vic_elec_daily |>
  ggplot(aes(x = Temperature, y = Demand, colour = Day_Type)) +
  geom_point(alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "U-shaped relationship between temperature and demand",
    x = "Maximum daily temperature (°C)",
    y = "Demand (GW)",
    colour = "Day type"
  )

ggplotly(p)
```

:::{.content-hidden unless-format="revealjs"}

##

:::

Two patterns are immediately clear:

:::{.incremental}

1. **U-shaped (quadratic) relationship** with temperature — both very hot and very cold days increase demand.
2. **Day type** matters: weekdays > weekends > holidays.

:::

:::{.fragment}

This gives us the model:

$$y_t = \beta_0 + \beta_1 T_t + \beta_2 T_t^2 + \beta_3 D_{\text{Weekday}} + \eta_t, \quad \eta_t \sim \text{ARIMA}$$

:::

:::{.content-hidden unless-format="revealjs"}

## Fitting the Model {auto-animate="true"}

```{r}
#| label: elec-fit-1
#| eval: false

vic_elec_daily |>
  model(
    ARIMA(Demand ~ Temperature)
  )
```

## Fitting the Model {auto-animate="true"}

```{r}
#| label: elec-fit-2
#| eval: false

vic_elec_daily |>
  model(
    ARIMA(Demand ~ Temperature + I(Temperature^2))
  )
```

## Fitting the Model {auto-animate="true"}

```{r}
#| label: elec-fit-3
#| eval: false

vic_elec_daily |>
  model(
    ARIMA(Demand ~ Temperature + I(Temperature^2) +
            (Day_Type == "Weekday"))
  )
```

:::

:::{.content-hidden unless-format="revealjs"}

## Fitting the Model {auto-animate="true"}

:::

:::{.content-visible unless-format="revealjs"}

### Fitting the Model

:::

```{r}
#| label: elec-fit-full

vic_elec_fit <- vic_elec_daily |>
  model(
    ARIMA(Demand ~ Temperature + I(Temperature^2) +  #<1>
            (Day_Type == "Weekday"))                  #<2>
  )

report(vic_elec_fit)
```

1. `I(Temperature^2)` adds the quadratic term. The `I()` wrapper is required to protect arithmetic inside the formula.
2. A logical expression that becomes a binary indicator: `TRUE` (1) for weekdays, `FALSE` (0) otherwise.

:::{.content-hidden unless-format="revealjs"}

## Diagnostics & Scenario Forecast

:::

:::{.content-visible unless-format="revealjs"}

### Diagnostics & Scenario Forecast

:::

```{r}
#| label: elec-diagnostics

vic_elec_fit |> gg_tsresiduals()
```

```{r}
#| label: elec-ljung

vic_elec_fit |>
  augment() |>
  features(.innov, ljung_box, dof = 8, lag = 14)
```

:::{.callout-note appearance="simple"}
## Heteroscedastic residuals

Variance is higher during January–February (Australian summer peak). Point estimates remain valid, but prediction intervals may be too narrow for extreme temperature periods.
:::

:::{.panel-tabset}

### Conservative (26°C)

```{r}
#| label: elec-forecast-conservative

vic_elec_future_26 <- new_data(vic_elec_daily, 14) |>
  mutate(
    Temperature = 26,
    Holiday     = c(TRUE, rep(FALSE, 13)),
    Day_Type    = case_when(
      Holiday              ~ "Holiday",
      wday(Date) %in% 2:6  ~ "Weekday",
      TRUE                 ~ "Weekend"
    )
  )

forecast(vic_elec_fit, vic_elec_future_26) |>
  autoplot(vic_elec_daily) +
  labs(title = "Scenario: constant 26°C",
       y = "Electricity demand (GW)")
```

### Heatwave

```{r}
#| label: elec-forecast-heatwave

vic_elec_future_hw <- new_data(vic_elec_daily, 14) |>
  mutate(
    Temperature = c(seq(31, 34), 33, rep(34, 3), rep(33, 6)), #<1>
    Holiday     = c(TRUE, rep(FALSE, 13)),
    Day_Type    = case_when(
      Holiday              ~ "Holiday",
      wday(Date) %in% 2:6  ~ "Weekday",
      TRUE                 ~ "Weekend"
    )
  )

forecast(vic_elec_fit, vic_elec_future_hw) |>
  autoplot(vic_elec_daily) +
  labs(title = "Scenario: rising temperatures (heatwave)",
       y = "Electricity demand (GW)")
```

1. A realistic meteorological forecast — considerably hotter than 26°C.

:::

# Harmonic Regression

:::{.content-hidden unless-format="revealjs"}

## A Problem with `vic_elec_daily`

:::

The electricity demand example above used **daily** data. Demand has an obvious **annual** seasonal pattern ($m = 365$). If we wanted to capture it with dummy variables, we would need 364 of them.

:::{.fragment}

That is impractical. The same problem appears with:

| Data frequency | Seasonal period |
|:---|:---:|
| Daily (annual cycle) | $m = 365$ |
| Weekly | $m = 52$ |
| Half-hourly (daily cycle) | $m = 48$ |

:::

:::{.fragment}

:::{.callout-tip appearance="simple"}
## Recall from Session 3.2

We first saw Fourier terms mentioned as a footnote in `regression.qmd` — they were promised as the solution for large $m$. This is where we deliver on that promise.
:::

:::

:::{.content-hidden unless-format="revealjs"}

## Fourier Terms

:::

:::{.content-visible unless-format="revealjs"}

### Fourier Terms

:::

Any periodic pattern can be approximated by a sum of sine and cosine waves:

$$S_K(t) = \sum_{k=1}^{K} \left[ a_k \cos\!\left(\frac{2\pi k t}{m}\right) + b_k \sin\!\left(\frac{2\pi k t}{m}\right) \right]$$

- $m$ — seasonal period
- $K$ — number of Fourier pairs (controls flexibility)
- $K = m/2$ — equivalent to $m-1$ seasonal dummies

:::{.fragment}

:::{.callout-important appearance="simple"}
## The key parameter: $K$

- **Small $K$** → smooth, slowly-varying seasonal pattern
- **Large $K$** → complex, flexible seasonal pattern (but more parameters)
- **$K = m/2$** → identical to using $m - 1$ seasonal dummies

Choose $K$ by minimizing AIC$_c$.
:::

:::

:::{.content-hidden unless-format="revealjs"}

## Advantages and Limitations

:::

:::{.content-visible unless-format="revealjs"}

### Advantages and Limitations

:::

::: {.columns}

::: {.column width="50%"}

**Advantages over seasonal dummies:**

:::{.incremental}

- Works for **any** $m$, no matter how large
- Handles **multiple seasonalities** simultaneously
- Smooth patterns with small $K$ → fewer parameters
- Short-run dynamics handled by ARIMA errors

:::

:::

::: {.column width="50%"}

**Limitations:**

:::{.incremental}

- Seasonal pattern assumed **fixed** over time
- Requires choosing $K$
- Less intuitive than dummies for small $m$

:::

:::

:::

# Example: Australian Café Spending

:::{.content-hidden unless-format="revealjs"}

## The Data

:::

Monthly retail turnover for cafés and restaurants in Australia (2004–2018). A clean example of a monthly series with strong, stable seasonality — ideal for comparing values of $K$:

```{r}
#| label: cafe-data

aus_cafe <- aus_retail |>
  filter(
    Industry == "Cafes, restaurants and takeaway food services",
    year(Month) %in% 2004:2018
  ) |>
  summarise(Turnover = sum(Turnover))

aus_cafe |> autoplot(Turnover) +
  labs(title = "Australian café & restaurant turnover",
       y = "AUD (millions)", x = NULL)
```

:::{.content-hidden unless-format="revealjs"}

## Fitting Multiple Values of K {auto-animate="true"}

```{r}
#| label: cafe-fit-1
#| eval: false

aus_cafe |>
  model(
    `K = 1` = ARIMA(log(Turnover) ~ fourier(K = 1) + PDQ(0, 0, 0))
  )
```

## Fitting Multiple Values of K {auto-animate="true"}

```{r}
#| label: cafe-fit-2
#| eval: false

aus_cafe |>
  model(
    `K = 1` = ARIMA(log(Turnover) ~ fourier(K = 1) + PDQ(0, 0, 0)),
    `K = 2` = ARIMA(log(Turnover) ~ fourier(K = 2) + PDQ(0, 0, 0)),
    `K = 3` = ARIMA(log(Turnover) ~ fourier(K = 3) + PDQ(0, 0, 0))
  )
```

:::

:::{.content-hidden unless-format="revealjs"}

## Fitting Multiple Values of K {auto-animate="true"}

:::

:::{.content-visible unless-format="revealjs"}

### Fitting Multiple Values of K

:::

```{r}
#| label: cafe-fit-full
#| message: false

cafe_fit <- aus_cafe |>
  model(
    `K = 1` = ARIMA(log(Turnover) ~ fourier(K = 1) + PDQ(0, 0, 0)), #<1>
    `K = 2` = ARIMA(log(Turnover) ~ fourier(K = 2) + PDQ(0, 0, 0)),
    `K = 3` = ARIMA(log(Turnover) ~ fourier(K = 3) + PDQ(0, 0, 0)),
    `K = 4` = ARIMA(log(Turnover) ~ fourier(K = 4) + PDQ(0, 0, 0)),
    `K = 5` = ARIMA(log(Turnover) ~ fourier(K = 5) + PDQ(0, 0, 0)),
    `K = 6` = ARIMA(log(Turnover) ~ fourier(K = 6) + PDQ(0, 0, 0))  #<2>
  )

glance(cafe_fit) |>
  select(.model, AICc, BIC) |>
  arrange(AICc)
```

1. `PDQ(0,0,0)` suppresses seasonal ARIMA terms — Fourier handles the seasonality entirely.
2. For monthly data ($m = 12$), $K = 6$ is exactly equivalent to using 11 seasonal dummies.

:::{.content-hidden unless-format="revealjs"}

## Comparing Forecasts Across K

:::

:::{.content-visible unless-format="revealjs"}

### Comparing Forecasts Across K

:::

```{r}
#| label: cafe-forecast-comparison

cafe_fit |>
  forecast(h = "2 years") |>
  autoplot(aus_cafe, level = NULL) +
  facet_wrap(~ .model, ncol = 2) +
  labs(
    title = "Effect of K on seasonal shape",
    y = "Turnover (AUD millions)", x = NULL
  ) +
  guides(colour = "none")
```

:::{.callout-note appearance="simple"}
## What to look for

- **K = 1–2**: Very smooth, nearly sinusoidal — misses the multi-peak retail calendar
- **K = 3–4**: Usually the AIC$_c$ winner; captures the main features of Australian retail seasonality
- **K = 6**: Reproduces seasonal dummies exactly — can overfit the historical pattern
:::

:::{.content-hidden unless-format="revealjs"}

## Multiple Seasonalities

:::

:::{.content-visible unless-format="revealjs"}

### Multiple Seasonalities

:::

The real power of Fourier terms appears with **nested seasonal periods**. Half-hourly electricity demand, for instance, has three simultaneous cycles:

:::{.incremental}

- **Daily** cycle ($m = 48$ half-hours)
- **Weekly** cycle ($m = 336$ half-hours)
- **Annual** cycle ($m \approx 17{,}520$ half-hours)

:::

:::{.fragment}

`fable`'s `fourier()` handles this with the `period` argument:

```{r}
#| eval: false

TSLM(demand ~ mean_temp + I(mean_temp^2) + day_type +
       fourier(period = "day",  K = 4) +   #<1>
       fourier(period = "week", K = 3) +   #<2>
       fourier(period = "year", K = 6))    #<3>
```

1. Within-day rhythm: 24-hour cycle.
2. Within-week rhythm: weekday/weekend pattern.
3. Annual seasonal pattern.

:::

:::{.callout-note appearance="simple"}
## Coming up in Module 4

In **4.1 Complex Seasonality** we will fit exactly this kind of model on `vic_elec` aggregated by hour — three nested periods, real data, with all the challenges that come with it. What you've seen here is the foundation.
:::

:::{.callout-warning appearance="simple"}
## Computation time

`ARIMA()` with multiple Fourier terms on thousands of rows can take several minutes. Start with `TSLM()` to explore values of $K$, then switch to `ARIMA()` once you have settled on a configuration.
:::

# Summary

:::{.content-hidden unless-format="revealjs"}

##

:::

:::{.incremental}

- The TSLM from Session 3.2 had one remaining weakness: **autocorrelated residuals**. Dynamic regression fixes this by letting the error term follow an ARIMA process instead of assuming white noise.

- The model has two error terms: regression errors $\eta_t$ (modeled by ARIMA) and innovations $\varepsilon_t$ (white noise). **Diagnose using $\hat{\varepsilon}_t$** — the innovations, not the regression residuals.

- **All variables must be stationary.** If differencing is needed, apply it to all simultaneously using `pdq(d = 1)`.

- Forecasting still requires **future values of the predictors** — the same ex-ante / ex-post / scenario strategies from Session 3.2 apply.

- **Harmonic regression** extends the framework to large or multiple seasonal periods using Fourier terms. Choose $K$ by minimizing AIC$_c$.

:::

:::{.fragment}

:::{.callout-tip appearance="simple"}
## The full Module 3 picture

| Session | Model | What was added |
|:---|:---|:---|
| 3.1 | Practical issues | Transformations, outliers, cross-validation |
| 3.2 | TSLM | External predictors; $\varepsilon_t$ white noise (required) |
| **3.3** | **ARIMA(y ~ x)** | **Same predictors; $\eta_t \sim$ ARIMA (relaxed)** |
| 3.4 | Prophet | Automatic changepoints, complex seasonality |

:::

:::
