---
title: "Model Diagnostics and Advanced Forecasting"
format:
  html:
    other-links:
      - text: FPP3 Chapter 5
        href: https://otexts.com/fpp3/toolbox.html
  revealjs:
    output-file: model_diagnostics_pres.html
freeze: auto
---

```{r}
#| label: setup
#| message: false
#| include: false

library(tidyverse)
library(fpp3)

gas_train <- aus_production |>
  filter_index(. ~ "2005 Q4")

gas_test <- aus_production |>
  filter_index("2006 Q1" ~ .)

gas_fit <- gas_train |>
  model(
    mean   = MEAN(Gas),
    naive  = NAIVE(Gas),
    snaive = SNAIVE(Gas),
    drift  = RW(Gas ~ drift())
  )

gas_fc <- gas_fit |>
  forecast(h = nrow(gas_test))
```

# Prediction intervals

::: {.content-hidden unless-format="revealjs"}
## Prediction intervals
:::

A point forecast tells you *where* the series is expected to go.
A **prediction interval** tells you *how confident* you should be.

::: {.fragment}

$$\hat{y}_{T+h|T} \pm c \cdot \hat{\sigma}_h$$

where $c$ depends on the coverage level (1.96 for 95%) and $\hat{\sigma}_h$ is the estimated forecast standard deviation at horizon $h$.

:::

::: {.fragment}

::: {.callout-important appearance="simple"}
## The interval is not optional

A point forecast without an interval is like a weather forecast without a probability of rain — it gives the illusion of certainty.
Uncertainty is part of the forecast.
:::

:::

::: {.content-hidden unless-format="revealjs"}
## One-step vs. multi-step
:::

::: {.content-visible unless-format="revealjs"}
### One-step vs. multi-step
:::

$\hat{\sigma}_h$ grows with the horizon $h$ — errors accumulate as we forecast further ahead.

For the benchmark methods:

| Method | $\hat{\sigma}_h$ |
|--------|-----------------|
| NAIVE  | $\hat{\sigma}\sqrt{h}$ |
| Drift  | $\hat{\sigma}\sqrt{h\left(1 + \frac{h}{T}\right)}$ |
| SNAIVE | $\hat{\sigma}\sqrt{\lfloor(h-1)/m\rfloor + 1}$ |
| MEAN   | $\hat{\sigma}\sqrt{1 + 1/T}$ |

The Mean interval barely widens; NAIVE and Drift intervals widen continuously.

::: {.content-hidden unless-format="revealjs"}
## Reading intervals with `hilo()`
:::

::: {.content-visible unless-format="revealjs"}
### Reading intervals with `hilo()`
:::

`autoplot()` shows intervals visually. `hilo()` extracts the numeric bounds:

::: {.fragment}

```{r}
#| label: hilo-example
#| df-print: tibble

gas_fc |>
  filter(.model == "snaive") |>
  hilo(level = 95) |>
  select(Quarter, .mean, `95%`) |>
  slice_head(n = 6)
```

:::

::: {.content-hidden unless-format="revealjs"}
## The problem with Gas
:::

::: {.content-visible unless-format="revealjs"}
### The problem with Gas
:::

Look at what SNAIVE and Drift produce for Gas:

::: {.fragment}

```{r}
#| label: gas-pi-raw
#| code-fold: true

gas_fc |>
  filter(.model %in% c("snaive", "drift")) |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = 95
  ) +
  facet_wrap(~ .model, nrow = 2) +
  labs(
    title = "Gas — benchmark forecasts with 95% prediction intervals",
    y = "Petajoules", x = NULL
  )
```

:::

::: {.fragment}

Both models produce intervals of **constant width** — they assume forecast variance stays the same across all periods. But the series shows growing variance: the swings in the 2000s are far larger than those in the 1970s. The intervals are miscalibrated.

:::

# Forecasting with transformations

::: {.content-hidden unless-format="revealjs"}
## The residuals tell the story
:::

::: {.content-visible unless-format="revealjs"}
### The residuals tell the story
:::

From the residual diagnostics we ran in [1.3](../03_fcst/forecasting.qmd), SNAIVE on Gas showed non-constant variance — larger residuals in recent decades, smaller in earlier ones. That is the signature of **multiplicative** behavior: the seasonal swings are proportional to the level of the series.

::: {.fragment}

```{r}
#| label: gas-resid-reminder
#| code-fold: true

gas_fit |>
  select(snaive) |>
  gg_tsresiduals() +
  labs(title = "Gas — SNAIVE residuals")
```

:::

::: {.fragment}

As we saw in [1.2](../02_ts_dcmp/ts_dcmp.qmd), a Box-Cox transformation stabilizes the variance. Applied here, the model fits in the transformed scale and `fable` back-transforms the forecasts automatically.

:::

::: {.content-hidden unless-format="revealjs"}
## Choosing $\lambda$
:::

::: {.content-visible unless-format="revealjs"}
### Choosing $\lambda$
:::

```{r}
#| label: guerrero

lambda <- aus_production |>
  features(Gas, features = guerrero) |>  #<1>
  pull(lambda_guerrero)                  #<2>

lambda
```

1. `guerrero` is a named feature function in `feasts` — pass it directly to `features()`.
2. Extract the scalar value so it can be reused in model specs.

::: {.content-hidden unless-format="revealjs"}
## Fitting models with Box-Cox
:::

::: {.content-visible unless-format="revealjs"}
### Fitting with Box-Cox
:::

Wrap the response variable in `box_cox()` inside the model spec:

::: {.fragment}

```{r}
#| label: gas-fit-bc

gas_fit_bc <- gas_train |>
  model(
    snaive_bc = SNAIVE(box_cox(Gas, lambda)),       #<1>
    drift_bc  = RW(box_cox(Gas, lambda) ~ drift())  #<2>
  )
```

1. Box-Cox applied inside the spec — the model sees stabilized variance.
2. Same transformation for Drift. Both use the same `lambda` estimated above.

:::

::: {.fragment}

```{r}
#| label: gas-fc-bc

gas_fc_bc <- gas_fit_bc |>
  forecast(h = nrow(gas_test))
```

:::

::: {.content-hidden unless-format="revealjs"}
## Intervals before and after transformation
:::

::: {.content-visible unless-format="revealjs"}
### Intervals before and after
:::

The intervals are computed in the transformed scale and then back-transformed. Because $e^x$ is a convex function, the back-transformation stretches the upper tail more than the lower one — the resulting intervals in the original scale are **asymmetric**.

::: {.panel-tabset}

## Without transformation

```{r}
#| label: gas-pi-no-bc
#| echo: false

gas_fc |>
  filter(.model %in% c("snaive", "drift")) |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = 95
  ) +
  facet_wrap(~ .model, nrow = 2) +
  labs(
    title = "Gas — no transformation",
    y = "Petajoules", x = NULL
  )
```

## With Box-Cox

```{r}
#| label: gas-pi-bc
#| echo: false

gas_fc_bc |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = 95
  ) +
  facet_wrap(~ .model, nrow = 2) +
  labs(
    title = paste0("Gas — Box-Cox (λ = ", round(lambda, 3), ")"),
    y = "Petajoules", x = NULL
  )
```

:::

::: {.content-hidden unless-format="revealjs"}
## Accuracy after transformation
:::

::: {.content-visible unless-format="revealjs"}
### Accuracy after transformation
:::

Does the transformation improve forecast accuracy?

```{r}
#| label: gas-accuracy-bc

bind_rows(
  gas_fc |>
    filter(.model %in% c("snaive", "drift")) |>
    accuracy(aus_production),
  gas_fc_bc |>
    accuracy(aus_production)
) |>
  select(.model, RMSE, MAE, MASE, RMSSE) |>
  arrange(RMSSE)
```

::: {.fragment}

::: {.callout-note appearance="simple"}
MASE and RMSSE are computed in the **original scale** regardless of the transformation — `fable` back-transforms before measuring error. This makes accuracy metrics directly comparable across transformed and untransformed models.
:::

:::


# Bias adjustment

::: {.content-hidden unless-format="revealjs"}
## Mean ≠ median after back-transformation
:::

::: {.content-visible unless-format="revealjs"}
### Mean ≠ median after back-transformation
:::

Back-transforming a forecast recovers the **median** of the forecast distribution in the original scale — not the mean.

::: {.fragment}

For a log transformation, if $w_{T+h} \sim \mathcal{N}(\mu, \sigma_h^2)$, then in the original scale:

$$\text{median}(\hat{y}_{T+h}) = e^{\mu} \qquad \text{mean}(\hat{y}_{T+h}) = e^{\mu + \sigma_h^2/2}$$

:::

::: {.fragment}

The median is invariant under monotone transformations — the 50th percentile in log scale maps exactly to the 50th percentile in the original scale. The mean is not: because $e^x$ is convex, Jensen's inequality gives $E[e^W] > e^{E[W]}$, and the gap grows with $\sigma_h^2$.

:::

::: {.content-hidden unless-format="revealjs"}
## The general Box-Cox case
:::

::: {.content-visible unless-format="revealjs"}
### The general Box-Cox case
:::

For a general Box-Cox transformation, the bias-adjusted back-transformation is:

$$y_t = \begin{cases} \exp(w_t)\left[1 + \dfrac{\sigma_h^2}{2}\right] & \text{if } \lambda = 0 \\[10pt] (\lambda w_t + 1)^{1/\lambda}\left[1 + \dfrac{\sigma_h^2(1-\lambda)}{2(\lambda w_t+1)^{2}}\right] & \text{otherwise} \end{cases}$$

`fable` applies this adjustment **automatically by default**.

::: {.content-hidden unless-format="revealjs"}
## Seeing the difference
:::

::: {.content-visible unless-format="revealjs"}
### Seeing the difference
:::

The `eggs` series from `prices` makes the difference visible — it is short and noisy, so $\sigma_h^2$ grows quickly and the gap between mean and median becomes noticeable:

::: {.fragment}

```{r}
#| label: eggs-setup
#| code-fold: true

eggs_tsb <- prices |>
  filter(!is.na(eggs))                              #<1>

eggs_fit <- eggs_tsb |>
  model(rw = RW(log(eggs) ~ drift()))

eggs_fc <- eggs_fit |>
  forecast(h = 50) |>
  mutate(.median = median(eggs))                    #<2>
```

1. `prices` is a native `fpp3` dataset. Removing `NA`s before fitting avoids gaps in the series.
2. `median()` extracts the distributional median from each forecast distribution — different from `.mean`, which already contains the bias-adjusted mean.

:::

::: {.fragment}

```{r}
#| label: eggs-plot
#| code-fold: true

eggs_fc |>
  autoplot(eggs_tsb, level = 80) +
  geom_line(
    data = eggs_fc,
    aes(y = .median),
    linetype = "dashed",
    color = "steelblue"
  ) +
  labs(
    title = "Egg prices — Drift with log transformation",
    subtitle = "Solid line: mean (bias-adjusted)  ·  Dashed: median",
    y = "USD per dozen",
    x = NULL
  )
```

:::

::: {.content-visible unless-format="revealjs"}

::: {.callout-note appearance="simple"}
## When does bias adjustment matter most?

For a single series the difference is usually small. It becomes important when **aggregating forecasts** — summing store-level forecasts into a regional total, for example. In that case, using medians instead of means introduces a systematic downward bias in the aggregate. We will revisit this in Module 4 when we cover hierarchical forecasting.
:::

:::


# Bootstrap prediction intervals

::: {.content-hidden unless-format="revealjs"}
## When normality fails
:::

::: {.content-visible unless-format="revealjs"}
### When normality fails
:::

All the prediction intervals above assume that forecast errors are **normally distributed**. When residuals are skewed or heavy-tailed, this assumption breaks down and the intervals are poorly calibrated.

::: {.fragment}

The key observation is that we can always write:

$$y_t = \hat{y}_{t|t-1} + e_t$$

so a future observation can be simulated as:

$$y_{T+1} = \hat{y}_{T+1|T} + e_{T+1}$$

Since the errors are uncorrelated, we can substitute $e_{T+1}$ with a **randomly resampled** historical residual. Repeating this for $y_{T+2}, y_{T+3}, \ldots$ gives one possible future path. Do it thousands of times and you have a full distribution of futures — no normality required.

:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg"}
:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}
:::

::: {.content-visible unless-format="revealjs"}
### Many possible futures
:::

Here are five possible futures for Google's closing price, each built by resampling historical residuals step by step:

```{r}
#| label: google-generate
#| include: false

google_2015 <- gafa_stock |> 
  filter(Symbol == "GOOG") |> 
  mutate(day = row_number()) |> 
  update_tsibble(index = day, regular = TRUE) |> 
  filter(year(Date) == 2015)

google_fit <- google_2015 |>
  model(NAIVE(Close))

google_sim <- google_fit |>
  generate(h = 30, times = 5, bootstrap = TRUE, seed = 123)
```

```{r}
#| label: google-sim-plot
#| echo: false

google_2015 |>
  ggplot(aes(x = day)) +
  geom_line(aes(y = Close)) +
  geom_line(
    data  = google_sim,
    aes(y = .sim, color = as.factor(.rep))
  ) +
  labs(
    title = "Google closing stock price — 5 simulated futures",
    y = "USD", x = NULL, color = "Path"
  ) +
  theme(legend.position = "bottom")
```

::: {.content-hidden unless-format="revealjs"}
## From paths to intervals
:::

::: {.content-visible unless-format="revealjs"}
### From paths to intervals
:::

Running this thousands of times and taking percentiles of the simulated paths gives the **bootstrap prediction interval**. `forecast()` does this in one step:

::: {.fragment}

```{r}
#| label: google-bootstrap

google_fc_boot <- google_fit |>
  forecast(
    h         = 30,
    bootstrap = TRUE,   #<1>
    times     = 1000    #<2>
  )
```

1. `bootstrap = TRUE` switches from the analytical normal interval to the resampling-based one.
2. Number of simulated paths — more gives smoother intervals at higher computational cost.

:::

::: {.fragment}

```{r}
#| label: google-bootstrap-plot
#| code-fold: true

google_fc_boot |>
  autoplot(google_2015, level = 95) +
  labs(
    title = "Google closing stock price — bootstrap 95% interval",
    y = "USD", x = NULL
  )
```

:::

::: {.fragment}

::: {.callout-tip appearance="simple"}
## When is bootstrap worth it?

Use bootstrap intervals when `gg_tsresiduals()` shows clearly non-normal residuals — heavy tails or marked skew. For series where residuals are roughly symmetric, the additional computational cost rarely changes the conclusion.
:::

:::

# Using STL Decomposition for forecasting using `decomposition_model()`

::: {.content-hidden unless-format="revealjs"}
## The idea
:::

::: {.content-visible unless-format="revealjs"}
### The idea
:::

STL (covered in [1.2](../02_ts_dcmp/ts_dcmp.qmd)) decomposes a series into trend-cycle, seasonal, and remainder components. The benchmarks (covered in [1.3](../03_fcst/forecasting.qmd)) each handle one aspect of the series well.

::: {.fragment}

`decomposition_model()` combines them into a pipeline:

1. Apply STL to separate the components
2. Forecast the **seasonally adjusted series** with one model
3. Re-seasonalize: add the seasonal component forecast back

:::

::: {.fragment}

::: {.callout-note appearance="simple"}
The reconstruction is additive in the decomposition scale. If the STL was applied to a transformed series, the reconstruction happens before back-transformation.
:::

:::

::: {.content-hidden unless-format="revealjs"}

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-1
#| eval: false

stl_spec <- decomposition_model(

)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-2
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season()),

)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-3
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season()),
  SNAIVE(season_adjust),

)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-4
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season()),
  SNAIVE(season_adjust),
  RW(season_adjust ~ drift())
)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-5
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season()),
  SNAIVE(season_adjust),
  RW(season_adjust ~ drift())
)

gas_train |> model(stl_snaive_drift = stl_spec)
```

:::

::: {.content-visible unless-format="revealjs"}
### Syntax
:::

The spec is defined once as a standalone object and passed into `model()`:

```{r}
#| label: dcmp-fit

stl_spec <- decomposition_model(   #<1>
  STL(Gas ~ trend() + season()),   #<2>
  SNAIVE(season_adjust),           #<3>
  RW(season_adjust ~ drift())      #<4>
)

gas_fit_dcmp <- gas_train |>
  model(stl_snaive_drift = stl_spec)
```

1. `decomposition_model()` wraps a decomposition method and one or more component models into a single reusable spec.
2. STL decomposition — same syntax as in 1.2.
3. SNAIVE on `season_adjust` handles the seasonal component extracted by STL.
4. Drift on `season_adjust` handles the trend-cycle. `fable` reconstructs the final forecast by adding the two component forecasts.

::: {.content-hidden unless-format="revealjs"}

## With a transformation {auto-animate="true"}

```{r}
#| label: dcmp-animate-bc-1
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season()),
  SNAIVE(season_adjust),
  RW(season_adjust ~ drift())
)

gas_train |> model(stl_snaive_drift = stl_spec)
```

## With a transformation {auto-animate="true"}

```{r}
#| label: dcmp-animate-bc-2
#| eval: false

stl_spec_bc <- decomposition_model(
  STL(box_cox(Gas, lambda) ~ trend() + season()),
  SNAIVE(season_adjust),
  RW(season_adjust ~ drift())
)

gas_train |> model(stl_bc = stl_spec_bc)
```

:::

::: {.content-visible unless-format="revealjs"}
### With a transformation
:::

Box-Cox goes inside the STL spec — declared once, inherited everywhere:

```{r}
#| label: dcmp-fit-bc

stl_spec_bc <- decomposition_model(              #<1>
  STL(box_cox(Gas, lambda) ~ trend() + season()),
  SNAIVE(season_adjust),
  RW(season_adjust ~ drift())
)

gas_fit_dcmp_bc <- gas_train |>
  model(stl_bc = stl_spec_bc)
```

1. The only difference from `stl_spec` is `box_cox()` in the STL call. All component models and the final back-transformation inherit it automatically.

```{r}
#| label: dcmp-fc-both

gas_fc_dcmp <- bind_rows(
  gas_fit_dcmp    |> forecast(h = nrow(gas_test)),
  gas_fit_dcmp_bc |> forecast(h = nrow(gas_test))
)
```

::: {.content-hidden unless-format="revealjs"}
## Final comparison
:::

::: {.content-visible unless-format="revealjs"}
### Final comparison
:::

All models side by side:

```{r}
#| label: gas-final-comparison
#| echo: false

bind_rows(
  gas_fc      |> filter(.model %in% c("snaive", "drift")),
  gas_fc_bc,
  gas_fc_dcmp
) |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = NULL
  ) +
  labs(
    title = "Gas — all Module 1 models",
    y = "Petajoules", x = NULL
  )
```

::: {.content-hidden unless-format="revealjs"}
## Accuracy table
:::

::: {.content-visible unless-format="revealjs"}
### Accuracy table
:::

```{r}
#| label: gas-final-accuracy

gas_accu <- bind_rows(
  gas_fc      |> filter(.model %in% c("snaive", "drift")) |> accuracy(aus_production),
  gas_fc_bc   |> accuracy(aus_production),
  gas_fc_dcmp |> accuracy(aus_production)
) |>
  select(.model, RMSE, MAE, MASE, RMSSE) |>
  arrange(RMSSE)

gas_accu
```

::: {.fragment}

::: {.callout-important appearance="simple"}
## This is your Module 1 baseline

The best-performing model from this table is your benchmark for the rest of the course. Every model we build in Modules 2, 3, and 4 must beat this number to justify its additional complexity.
:::

:::

::: {.content-hidden unless-format="revealjs"}
## Refit on all data
:::

::: {.content-visible unless-format="revealjs"}
### Refit on all data
:::

Because the spec is a standalone object, refitting on the full dataset reuses it directly — no duplication:

```{r}
#| label: gas-final-refit

gas_final_fit <- aus_production |>
  model(stl_bc = stl_spec_bc)   #<1>

gas_final_fc <- gas_final_fit |>
  forecast(h = "2 years")

gas_final_fc |>
  autoplot(aus_production |> filter_index("2000 Q1" ~ .)) +
  labs(
    title = "Australian gas production — STL + Box-Cox forecast",
    subtitle = "Refit on full data · 2-year horizon",
    y = "Petajoules", x = NULL
  )
```

1. Same spec, full data — no need to rewrite the model definition.

::: {.fragment}

::: {.callout-note appearance="simple"}
## Module 1 complete

You now have a principled, validated baseline model. It handles trend, seasonality, and non-constant variance. Modules 2 onward will replace or enhance individual components — but the workflow stays the same: split, fit, diagnose, compare, refit.
:::

:::
