---
title: "Model Diagnostics and Advanced Forecasting"
format:
  html:
    other-links:
      - text: FPP3 Chapter 5
        href: https://otexts.com/fpp3/toolbox.html
  revealjs:
    output-file: model_diagnostics_pres.html
---

```{r}
#| label: setup
#| message: false
#| include: false

library(tidyverse)
library(fpp3)

gas_train <- aus_production |>
  filter_index(. ~ "2005 Q4")

gas_test <- aus_production |>
  filter_index("2006 Q1" ~ .)

gas_fit <- gas_train |>
  model(
    mean   = MEAN(Gas),
    naive  = NAIVE(Gas),
    snaive = SNAIVE(Gas),
    drift  = RW(Gas ~ drift())
  )

gas_fc <- gas_fit |>
  forecast(h = nrow(gas_test))

beer <- aus_production |>
  filter(year(Quarter) >= 1992) |>
  select(Quarter, Beer)

h_beer <- 4 * 4

beer_train <- beer |>
  slice_head(n = nrow(beer) - h_beer)

beer_fit <- beer_train |>
  model(
    mean   = MEAN(Beer),
    snaive = SNAIVE(Beer)
  )

beer_fc <- beer_fit |>
  forecast(h = h_beer)

source(here::here("R/narsil_theme.R"))
theme_set(theme_narsil())
```

# Prediction intervals

::: {.content-hidden unless-format="revealjs"}
## Why are these intervals so different?
:::

::: {.content-visible unless-format="revealjs"}
### Why are these intervals so different?
:::

```{r}
#| label: beer-pi-compare
#| code-fold: true
#| output: false

beer_pi_compare_p <- beer_fc |>
  autoplot(beer, level = 95) +        #<1>
  facet_wrap(~ .model, nrow = 1) +    #<2>
  labs(
    title = "Beer — same series, two benchmark models, 95% prediction intervals",
    y = "Megalitres", x = NULL
  )
```

1. `beer_fc` holds the forecasts of both models; `level = 95` draws only the 95% band.
2. One panel per model, side by side and sharing the y-axis, so the interval widths are directly comparable.

```{r}
#| label: beer-pi-compare-render
#| renderings: [light, dark]
#| echo: false

beer_pi_compare_p + theme_narsil()
beer_pi_compare_p + theme_narsil_dark()
```

::: {.fragment}

**Same data. Same coverage level. Why is one interval several times wider than the other?**

:::

::: {.content-hidden unless-format="revealjs"}
::: {.notes}
Let students answer before advancing. Expected answer: the MEAN model fits badly, so its residuals are large.
:::
:::

::: {.content-hidden unless-format="revealjs"}
## The interval width comes from the residuals
:::

::: {.content-visible unless-format="revealjs"}
### The interval width comes from the residuals
:::

The one-step residual standard deviation is

$$\hat{\sigma} = \sqrt{\frac{1}{T-K-M}\sum_{t=1}^{T} e_t^2}$$

where $K$ is the number of parameters estimated and $M$ the number of missing residuals at the start of the series (e.g. $m$ for SNAIVE).

::: {.fragment}

```{r}
#| label: beer-sigma

beer_fit |>
  augment() |>                                     #<1>
  as_tibble() |>
  group_by(.model) |>
  summarise(sigma_hat = sd(.innov, na.rm = TRUE))  #<2>
```

1. `augment()` returns fitted values and residuals for every model in the mable.
2. The standard deviation of the innovation residuals is the $\hat\sigma$ that every prediction interval is built on. `sd()` divides by $T-1$ rather than $T-K-M$; the difference is negligible here, and `fable` uses the exact formula internally.

:::

::: {.fragment}

Every formula on the next slides multiplies this single number by a horizon-dependent factor. A model that fits poorly produces wide intervals — the interval is honest about how little the model knows.

:::

::: {.content-hidden unless-format="revealjs"}
## Prediction intervals
:::

A point forecast tells you *where* the series is expected to go.
A **prediction interval** tells you *how confident* you should be.

::: {.fragment}

$$\hat{y}_{T+h|T} \pm c \cdot \hat{\sigma}_h$$

where $c$ depends on the coverage level (1.96 for 95%) and $\hat{\sigma}_h$ is the estimated forecast standard deviation at horizon $h$, built from the residual standard deviation $\hat\sigma$ we just computed.

:::

::: {.fragment}

::: {.callout-important appearance="simple"}
## The interval is not optional

A point forecast without an interval is like a weather forecast without a probability of rain — it gives the illusion of certainty.
Uncertainty is part of the forecast.
:::

:::

::: {.content-hidden unless-format="revealjs"}
## One-step vs. multi-step
:::

::: {.content-visible unless-format="revealjs"}
### One-step vs. multi-step
:::

$\hat{\sigma}_h$ grows with the horizon $h$ — errors accumulate as we forecast further ahead.

For the benchmark methods:

| Method | $\hat{\sigma}_h$ |
|--------|-----------------|
| $h = 1$ (all methods) | $\hat{\sigma}$ |
| NAIVE  | $\hat{\sigma}\sqrt{h}$ |
| Drift  | $\hat{\sigma}\sqrt{h\left(1 + \frac{h}{T}\right)}$ |
| SNAIVE | $\hat{\sigma}\sqrt{\lfloor(h-1)/m\rfloor + 1}$ |
| MEAN   | $\hat{\sigma}\sqrt{1 + 1/T}$ |

The Mean interval barely widens; NAIVE and Drift intervals widen continuously.

::: {.content-hidden unless-format="revealjs"}
## Reading intervals with `hilo()`
:::

::: {.content-visible unless-format="revealjs"}
### Reading intervals with `hilo()`
:::

`autoplot()` shows intervals visually. `hilo()` extracts the numeric bounds:

::: {.fragment}

```{r}
#| label: hilo-example
#| paged.print: false
#| code-line-numbers: "3"

gas_fc |>
  filter(.model == "snaive") |>
  hilo(level = 95) |>                #<1>
  select(Quarter, .mean, `95%`) |>   #<2>
  slice_head(n = 6)
```

1. `hilo()` converts each forecast distribution into a numeric interval at the requested level.
2. The interval lives in a new column named after the level; `.mean` is the point forecast.

:::

::: {.content-hidden unless-format="revealjs"}
## The problem with Gas
:::

::: {.content-visible unless-format="revealjs"}
### The problem with Gas
:::

Look at what SNAIVE and Drift produce for Gas:

::: {.fragment}

```{r}
#| label: gas-pi-raw
#| code-fold: true
#| output: false

gas_pi_raw_p <- gas_fc |>
  filter(.model %in% c("snaive", "drift")) |>   #<1>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),  #<2>
    level = 95
  ) +
  facet_wrap(~ .model, nrow = 2) +                #<3>
  labs(
    title = "Gas — benchmark forecasts with 95% prediction intervals",
    y = "Petajoules", x = NULL
  )
```

1. Keep only the two benchmarks worth comparing here.
2. Passing the full series (not just the training set) draws the test observations so we can see them against the intervals. Zooming in from 2000 makes the intervals readable.
3. One panel per model, stacked so the two intervals share the time axis.

```{r}
#| label: gas-pi-raw-render
#| renderings: [light, dark]
#| echo: false

gas_pi_raw_p + theme_narsil()
gas_pi_raw_p + theme_narsil_dark()
```

:::

::: {.content-visible unless-format="revealjs"}

Look at the SNAIVE panel: the observed peaks in the test set repeatedly fall **outside** the 95% interval. The interval is too narrow — and the reason is how $\hat\sigma$ was estimated. It is a single number computed from residuals across the whole training period (1956–2005), including decades when seasonal swings were a fraction of today's. Averaging small early residuals with large recent ones **understates current uncertainty**. The Drift panel has the opposite problem: a wide, symmetric interval that ignores the seasonal pattern entirely.

Neither model can express what the series clearly shows — variance that grows with the level. That is what a transformation fixes.

:::

::: {.content-hidden unless-format="revealjs"}

::: {.fragment}

- **SNAIVE:** observed peaks fall **outside** the 95% interval — too narrow.
- **Why:** one $\hat\sigma$ from all residuals, 1956–2005. Small early swings dilute the large recent ones.
- **Drift:** wide, symmetric, blind to seasonality.
- Neither can express variance that **grows with the level**. A transformation can.

:::

::: {.notes}
Look at the SNAIVE panel: the observed peaks in the test set repeatedly fall **outside** the 95% interval. The interval is too narrow — and the reason is how $\hat\sigma$ was estimated. It is a single number computed from residuals across the whole training period (1956–2005), including decades when seasonal swings were a fraction of today's. Averaging small early residuals with large recent ones **understates current uncertainty**. The Drift panel has the opposite problem: a wide, symmetric interval that ignores the seasonal pattern entirely.

Neither model can express what the series clearly shows — variance that grows with the level. That is what a transformation fixes.
:::

:::

# Forecasting with transformations

::: {.content-hidden unless-format="revealjs"}
## The residuals tell the story
:::

::: {.content-visible unless-format="revealjs"}
### The residuals tell the story
:::

From the residual diagnostics we ran in [1.3](../03_fcst/forecasting.qmd), SNAIVE on Gas showed non-constant variance — larger residuals in recent decades, smaller in earlier ones. That is the signature of **multiplicative** behavior: the seasonal swings are proportional to the level of the series.

::: {.fragment}

```{r}
#| label: gas-resid-reminder
#| code-fold: true
#| results: hide

theme_narsil()                                #<1>
gas_fit |>
  select(snaive) |>                           #<2>
  gg_tsresiduals() +
  labs(title = "Gas — SNAIVE residuals")
```

1. `gg_tsresiduals()` cannot use the light/dark render pattern, so the theme is reset explicitly first.
2. `gg_tsresiduals()` works on a single model: select one column of the mable before calling it.

:::

::: {.fragment}

As we saw in [1.2](../02_ts_dcmp/ts_dcmp.qmd), a Box-Cox transformation stabilizes the variance. Applied here, the model fits in the transformed scale and `fable` back-transforms the forecasts automatically.

:::

::: {.content-hidden unless-format="revealjs"}
## Choosing $\lambda$
:::

::: {.content-visible unless-format="revealjs"}
### Choosing $\lambda$
:::

```{r}
#| label: guerrero

lambda <- aus_production |>
  features(Gas, features = guerrero) |>  #<1>
  pull(lambda_guerrero)                  #<2>

lambda
```

1. `guerrero` is a named feature function in `feasts` — pass it directly to `features()`.
2. Extract the scalar value so it can be reused in model specs.

::: {.content-hidden unless-format="revealjs"}
## Fitting models with Box-Cox
:::

::: {.content-visible unless-format="revealjs"}
### Fitting with Box-Cox
:::

Wrap the response variable in `box_cox()` inside the model spec:

::: {.fragment}

```{r}
#| label: gas-fit-bc
#| code-line-numbers: "3,4"

gas_fit_bc <- gas_train |>
  model(
    snaive_bc = SNAIVE(box_cox(Gas, lambda)),       #<1>
    drift_bc  = RW(box_cox(Gas, lambda) ~ drift())  #<2>
  )
```

1. Box-Cox applied inside the spec — the model sees stabilized variance.
2. Same transformation for Drift. Both use the same `lambda` estimated above.

:::

::: {.fragment}

```{r}
#| label: gas-fc-bc

gas_fc_bc <- gas_fit_bc |>
  forecast(h = nrow(gas_test))   #<1>
```

1. Same horizon as the untransformed benchmarks. Nothing about the transformation appears here — see the next slide for why.

:::

::: {.content-hidden unless-format="revealjs"}
## Back to the original scale
:::

::: {.content-visible unless-format="revealjs"}
### Back to the original scale
:::

We fitted the model to $w_t = \text{box\_cox}(y_t, \lambda)$, so the forecasts and their intervals live in the transformed scale. To report them in the original units we apply the inverse transformation — the **back-transformation**:

$$y_t = \begin{cases} \exp(w_t) & \text{if } \lambda = 0 \\[6pt] \operatorname{sign}(\lambda w_t + 1)\,|\lambda w_t + 1|^{1/\lambda} & \text{otherwise} \end{cases}$$

::: {.fragment}

::: {.callout-note appearance="simple"}
`fable` does this for you. Because the transformation was declared inside the model formula, `forecast()` returns everything — point forecasts, intervals, the full distribution — already back-transformed. You never call `inv_box_cox()` yourself.
:::

:::

::: {.fragment}

Back-transforming is not free, though. It does two things to the forecast distribution, and both matter.

:::

::: {.content-hidden unless-format="revealjs"}
## Intervals before and after transformation
:::

::: {.content-visible unless-format="revealjs"}
### Intervals before and after
:::

The intervals are computed in the transformed scale and back-transformed with the inverse above. Because the inverse is a convex function, the back-transformation stretches the upper tail more than the lower one — the resulting intervals in the original scale are **asymmetric**.

::: {.panel-tabset}

## Without transformation

```{r}
#| label: gas-pi-no-bc-render
#| renderings: [light, dark]
#| echo: false

gas_pi_raw_p + theme_narsil()
gas_pi_raw_p + theme_narsil_dark()
```

## With Box-Cox

```{r}
#| label: gas-pi-bc
#| echo: false
#| output: false

gas_pi_bc_p <- gas_fc_bc |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = 95
  ) +
  facet_wrap(~ .model, nrow = 2) +
  labs(
    title = paste0("Gas — Box-Cox (λ = ", round(lambda, 3), ")"),
    y = "Petajoules", x = NULL
  )
```

```{r}
#| label: gas-pi-bc-render
#| renderings: [light, dark]
#| echo: false

gas_pi_bc_p + theme_narsil()
gas_pi_bc_p + theme_narsil_dark()
```

:::

::: {.content-hidden unless-format="revealjs"}
## Accuracy after transformation
:::

::: {.content-visible unless-format="revealjs"}
### Accuracy after transformation
:::

Does the transformation improve forecast accuracy?

```{r}
#| label: gas-accuracy-bc
#| code-fold: true

bind_rows(                                       #<1>
  gas_fc |>
    filter(.model %in% c("snaive", "drift")) |>
    accuracy(aus_production),                    #<2>
  gas_fc_bc |>
    accuracy(aus_production)
) |>
  select(.model, RMSE, MAE, MAPE, MASE, RMSSE) |>
  arrange(RMSSE)                                 #<3>
```

1. The two fables have different models, so stack their accuracy tables to compare them in one place.
2. `accuracy()` on a fable takes the **full** series and matches the test observations by index — never pass the test set alone.
3. Sort by the scaled error so the best model is at the top.

::: {.fragment}

::: {.callout-note appearance="simple"}
MASE and RMSSE are computed in the **original scale** regardless of the transformation — `fable` back-transforms before measuring error. This makes accuracy metrics directly comparable across transformed and untransformed models.
:::

:::


# Bias adjustment

::: {.content-hidden unless-format="revealjs"}
## Mean ≠ median after back-transformation
:::

::: {.content-visible unless-format="revealjs"}
### Mean ≠ median after back-transformation
:::

The second consequence of back-transforming: the point forecast you recover is the **median** of the forecast distribution in the original scale — not the mean. The back-transformation introduces a bias, and by default your forecasts follow the median.

::: {.fragment}

For a log transformation, if $w_{T+h} \sim \mathcal{N}(\mu, \sigma_h^2)$, then in the original scale:

$$\text{median}(\hat{y}_{T+h}) = e^{\mu} \qquad \text{mean}(\hat{y}_{T+h}) = e^{\mu + \sigma_h^2/2}$$

:::

::: {.fragment}

The median is invariant under monotone transformations — the 50th percentile in log scale maps exactly to the 50th percentile in the original scale. The mean is not: because $e^x$ is convex, Jensen's inequality gives $E[e^W] > e^{E[W]}$, and the gap grows with $\sigma_h^2$.

:::

::: {.content-visible unless-format="revealjs"}
::: {.callout-note collapse="true"}
## Why the median survives and the mean does not

A monotone increasing function $g$ preserves order: if $a < b$ then $g(a) < g(b)$. Percentiles are defined by order, so the value with 50% of the distribution below it in the $w$ scale maps to the value with 50% below it in the $y$ scale. Hence $\text{median}(g(W)) = g(\text{median}(W))$. The same holds for every percentile, which is why the prediction interval bounds can be back-transformed directly.

The mean is an average, not an order statistic. Averaging and then transforming is not the same as transforming and then averaging unless $g$ is linear. For convex $g$ — every Box-Cox inverse with $\lambda < 1$, including $\exp$ — Jensen's inequality gives $E[g(W)] \ge g(E[W])$, and the gap grows with the spread of $W$. Since $W$ is symmetric (normal), its mean equals its median, so $g(E[W]) = g(\text{median}(W)) = \text{median}(Y)$: the naive back-transform lands on the median and sits below the mean.
:::
:::

::: {.content-hidden unless-format="revealjs"}
::: {.notes}
A monotone increasing function $g$ preserves order: if $a < b$ then $g(a) < g(b)$. Percentiles are defined by order, so the value with 50% of the distribution below it in the $w$ scale maps to the value with 50% below it in the $y$ scale. Hence $\text{median}(g(W)) = g(\text{median}(W))$. The same holds for every percentile, which is why the prediction interval bounds can be back-transformed directly.

The mean is an average, not an order statistic. Averaging and then transforming is not the same as transforming and then averaging unless $g$ is linear. For convex $g$ — every Box-Cox inverse with $\lambda < 1$, including $\exp$ — Jensen's inequality gives $E[g(W)] \ge g(E[W])$, and the gap grows with the spread of $W$. Since $W$ is symmetric (normal), its mean equals its median, so $g(E[W]) = g(\text{median}(W)) = \text{median}(Y)$: the naive back-transform lands on the median and sits below the mean.
:::
:::

::: {.content-hidden unless-format="revealjs"}
## The general Box-Cox case
:::

::: {.content-visible unless-format="revealjs"}
### The general Box-Cox case
:::

For a general Box-Cox transformation, the bias-adjusted back-transformation is:

$$y_t = \begin{cases} \exp(w_t)\left[1 + \dfrac{\sigma_h^2}{2}\right] & \text{if } \lambda = 0 \\[10pt] (\lambda w_t + 1)^{1/\lambda}\left[1 + \dfrac{\sigma_h^2(1-\lambda)}{2(\lambda w_t+1)^{2}}\right] & \text{otherwise} \end{cases}$$

`fable` applies this adjustment **automatically by default**.

::: {.fragment}

::: {.callout-note appearance="simple"}
**This is an approximation.** The bracketed factor comes from a second-order Taylor expansion of the inverse transformation around $w_t$. For the log case the exact mean of a lognormal is $e^{\mu + \sigma_h^2/2}$, and $e^{\sigma_h^2/2} \approx 1 + \sigma_h^2/2$ when $\sigma_h^2$ is small — which is why the previous slide and this one agree to first order but not exactly. For the general Box-Cox case there is no closed form, so the approximation is what is used in practice.
:::

:::

::: {.content-hidden unless-format="revealjs"}
::: {.notes}
This is an approximation. The bracketed factor comes from a second-order Taylor expansion of the inverse transformation around $w_t$. For the log case the exact mean of a lognormal is $e^{\mu + \sigma_h^2/2}$, and $e^{\sigma_h^2/2} \approx 1 + \sigma_h^2/2$ when $\sigma_h^2$ is small — which is why the previous slide and this one agree to first order but not exactly. For the general Box-Cox case there is no closed form, so the approximation is what is used in practice.
:::
:::

::: {.content-hidden unless-format="revealjs"}
## Seeing the difference
:::

::: {.content-visible unless-format="revealjs"}
### Seeing the difference
:::

The `eggs` series from `prices` makes the difference visible — it is short and noisy, so $\sigma_h^2$ grows quickly and the gap between mean and median becomes noticeable:

::: {.fragment}

```{r}
#| label: eggs-setup
#| code-fold: true

eggs_tsb <- prices |>
  filter(!is.na(eggs))                              #<1>

eggs_fit <- eggs_tsb |>
  model(rw = RW(log(eggs) ~ drift()))

eggs_fc <- eggs_fit |>
  forecast(h = 50) |>
  mutate(.median = median(eggs))                    #<2>
```

1. `prices` is a native `fpp3` dataset. Removing `NA`s before fitting avoids gaps in the series.
2. `median()` extracts the distributional median from each forecast distribution — different from `.mean`, which already contains the bias-adjusted mean.

:::

::: {.fragment}

```{r}
#| label: eggs-plot
#| code-fold: true
#| output: false

eggs_plot_p <- eggs_fc |>
  autoplot(eggs_tsb, level = 80) +   #<1>
  geom_line(
    data = eggs_fc,
    aes(y = .median),                #<2>
    linetype = "dashed"
  ) +
  labs(
    title = "Egg prices — Drift with log transformation",
    subtitle = "Solid line: mean (bias-adjusted)  ·  Dashed: median",
    y = "USD per dozen",
    x = NULL
  )
```

1. `autoplot()` draws the bias-adjusted `.mean` as the point forecast, with an 80% interval.
2. Overlay the `.median` column computed above as a dashed line so the gap is visible.

```{r}
#| label: eggs-plot-render
#| renderings: [light, dark]
#| echo: false

eggs_plot_p + theme_narsil()
eggs_plot_p + theme_narsil_dark()
```

:::

::: {.content-visible unless-format="revealjs"}

::: {.callout-note appearance="simple"}
## When does bias adjustment matter most?

For a single series the difference is usually small. It becomes important when **aggregating forecasts** — summing store-level forecasts into a regional total, for example. In that case, using medians instead of means introduces a systematic downward bias in the aggregate. We will revisit this in Module 4 when we cover hierarchical forecasting.
:::

:::


# Bootstrap prediction intervals

::: {.content-hidden unless-format="revealjs"}
## When normality fails
:::

::: {.content-visible unless-format="revealjs"}
### When normality fails
:::

All the prediction intervals above assume that forecast errors are **normally distributed**. When residuals are skewed or heavy-tailed, this assumption breaks down and the intervals are poorly calibrated.

::: {.fragment}

**Bootstrapping** means resampling **with replacement** from the observed data and treating the empirical distribution as a stand-in for the unknown true distribution.

::: {.content-visible unless-format="revealjs"}
Here the observed data we resample are the model's historical residuals: if the model is correctly specified they are approximately independent draws from the error distribution, whatever shape it has.
:::

::: {.content-hidden unless-format="revealjs"}
::: {.notes}
Here the observed data we resample are the model's historical residuals: if the model is correctly specified they are approximately independent draws from the error distribution, whatever shape it has.
:::
:::

:::

::: {.fragment}

The key observation is that we can always write:

$$y_t = \hat{y}_{t|t-1} + e_t$$

so a future observation can be simulated as:

$$y_{T+1} = \hat{y}_{T+1|T} + e_{T+1}$$

Since the errors are uncorrelated, we can substitute $e_{T+1}$ with a **randomly resampled (with replacement)** historical residual. Repeating this for $y_{T+2}, y_{T+3}, \ldots$ gives one possible future path. Do it thousands of times and you have a full distribution of futures — no normality required.

:::

::: {.fragment}

::: {.callout-tip appearance="simple"}
The word *bootstrap* will return in Module 4 with a different object being resampled: there we bootstrap **entire series** (via STL and the remainder component) to build bagged forecasts. Here we bootstrap **residuals** to build prediction intervals. Same idea, different target.
:::

:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg"}
:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}
:::

```{r}
#| label: google-generate
#| include: false

google_2015 <- gafa_stock |>
  filter(Symbol == "GOOG") |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE) |>
  filter(year(Date) == 2015)

google_fit <- google_2015 |>
  model(NAIVE(Close))

google_sim <- google_fit |>
  generate(h = 30, times = 500, bootstrap = TRUE, seed = 123)

# .rep is a character column in fabletools, so filter on character labels.
# All three objects come from the same generate() call: the first 5 paths
# are identical across the 5 / 50 / 500 plots.
google_sim_5  <- google_sim |> filter(.rep %in% as.character(1:5))
google_sim_50 <- google_sim |> filter(.rep %in% as.character(1:50))
```

```{r}
#| label: google-fc-boot-hidden
#| include: false

# Computed here so the interval can be drawn on top of the simulated paths.
# The visible teaching chunk on the "From paths to intervals" slide re-runs it.
set.seed(123)
google_fc_boot <- google_fit |>
  forecast(h = 30, bootstrap = TRUE, times = 500)
```

```{r}
#| label: google-plot-futures
#| include: false

anduin <- narsil_pal("light")$palette[1]  # Gondor primary blue

plot_futures <- function(sim, title, alpha = 1, colour_by_rep = TRUE, legend = TRUE) {
  p <- google_2015 |>
    ggplot(aes(x = day)) +
    geom_line(aes(y = Close))
  if (colour_by_rep) {
    p <- p + geom_line(data = sim, aes(y = .sim, colour = as.factor(.rep)), alpha = alpha)
  } else {
    p <- p + geom_line(data = sim, aes(y = .sim, group = .rep), alpha = alpha, colour = anduin)
  }
  p + labs(title = title, y = "USD", x = NULL, colour = "Path") +
    theme(legend.position = if (legend) "bottom" else "none")
}

google_sim_5_p <- plot_futures(
  google_sim_5, "Google closing price — 5 simulated futures"
)

google_sim_50_p <- plot_futures(
  google_sim_50, "Google closing price — 50 simulated futures",
  alpha = 0.6, legend = FALSE
)

google_sim_500_p <- plot_futures(
  google_sim, "Google closing price — 500 simulated futures",
  alpha = 0.15, colour_by_rep = FALSE, legend = FALSE
)

google_pi95 <- google_fc_boot |>
  hilo(level = 95) |>
  unpack_hilo(`95%`) |>
  as_tibble()

google_sim_pi_p <- plot_futures(
  google_sim, "Google closing price — 500 futures and the 95% bootstrap interval",
  alpha = 0.15, colour_by_rep = FALSE, legend = FALSE
) +
  geom_line(data = google_pi95, aes(y = `95%_lower`), linetype = "dashed", linewidth = 0.9) +
  geom_line(data = google_pi95, aes(y = `95%_upper`), linetype = "dashed", linewidth = 0.9) +
  geom_line(data = google_pi95, aes(y = .mean), linewidth = 0.9)
```

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}

Five possible futures, each built by resampling a historical residual at every step:

```{r}
#| label: google-sim-5-render
#| renderings: [light, dark]
#| echo: false

google_sim_5_p + theme_narsil() + theme(legend.position = "bottom")
google_sim_5_p + theme_narsil_dark() + theme(legend.position = "bottom")
```

:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}

Fifty.

```{r}
#| label: google-sim-50-render
#| renderings: [light, dark]
#| echo: false

google_sim_50_p + theme_narsil() + theme(legend.position = "none")
google_sim_50_p + theme_narsil_dark() + theme(legend.position = "none")
```

:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}

Five hundred. A distribution is taking shape.

```{r}
#| label: google-sim-500-render
#| renderings: [light, dark]
#| echo: false

google_sim_500_p + theme_narsil() + theme(legend.position = "none")
google_sim_500_p + theme_narsil_dark() + theme(legend.position = "none")
```

:::

::: {.content-hidden unless-format="revealjs"}
## Many possible futures {auto-animate=true background-color="black" background-image="../00_intro/dr_strange_math.jpg" background-opacity="0.3"}

Take the 2.5th and 97.5th percentiles at each horizon and you have the bootstrap prediction interval.

```{r}
#| label: google-sim-pi-render
#| renderings: [light, dark]
#| echo: false

google_sim_pi_p + theme_narsil() + theme(legend.position = "none")
google_sim_pi_p + theme_narsil_dark() + theme(legend.position = "none")
```

:::

::: {.content-visible unless-format="revealjs"}
### Many possible futures

Each path is built by resampling a historical residual at every step. Flip through the tabs to watch a distribution emerge from individual paths.

::: {.panel-tabset}

## 5 paths

Five possible futures for Google's closing price.

```{r}
#| label: google-sim-5-html-render
#| renderings: [light, dark]
#| echo: false

google_sim_5_p + theme_narsil() + theme(legend.position = "bottom")
google_sim_5_p + theme_narsil_dark() + theme(legend.position = "bottom")
```

## 50 paths

Fifty paths. The first five are the same as in the previous tab.

```{r}
#| label: google-sim-50-html-render
#| renderings: [light, dark]
#| echo: false

google_sim_50_p + theme_narsil() + theme(legend.position = "none")
google_sim_50_p + theme_narsil_dark() + theme(legend.position = "none")
```

## 500 paths

Five hundred paths. A distribution is taking shape.

```{r}
#| label: google-sim-500-html-render
#| renderings: [light, dark]
#| echo: false

google_sim_500_p + theme_narsil() + theme(legend.position = "none")
google_sim_500_p + theme_narsil_dark() + theme(legend.position = "none")
```

## 500 paths + interval

Take the 2.5th and 97.5th percentiles at each horizon and you have the bootstrap prediction interval (dashed) around the bootstrap mean (solid).

```{r}
#| label: google-sim-pi-html-render
#| renderings: [light, dark]
#| echo: false

google_sim_pi_p + theme_narsil() + theme(legend.position = "none")
google_sim_pi_p + theme_narsil_dark() + theme(legend.position = "none")
```

:::

:::

::: {.content-hidden unless-format="revealjs"}
## From paths to intervals
:::

::: {.content-visible unless-format="revealjs"}
### From paths to intervals
:::

Running this hundreds or thousands of times and taking percentiles of the simulated paths gives the **bootstrap prediction interval**. `forecast()` does this in one step:

::: {.fragment}

```{r}
#| label: google-bootstrap
#| code-line-numbers: "|4,5|6"

google_fc_boot <- google_fit |>
  forecast(
    h         = 30,
    bootstrap = TRUE,   #<1>
    times     = 500,    #<2>
    seed      = 123     #<3>
  )
```

1. `bootstrap = TRUE` switches from the analytical normal interval to the resampling-based one.
2. Number of simulated paths — 500 here, matching the paths drawn on the previous slides. More gives smoother intervals at higher computational cost.
3. Fixed seed so the interval is reproducible across renders.

:::

::: {.fragment}

```{r}
#| label: google-bootstrap-plot
#| code-fold: true
#| output: false

google_bootstrap_plot_p <- google_fc_boot |>
  autoplot(google_2015, level = 95) +   #<1>
  labs(
    title = "Google closing stock price — bootstrap 95% interval",
    y = "USD", x = NULL
  )
```

1. A bootstrap fable plots exactly like a normal one — `autoplot()` reads the interval from the sample distribution instead of a normal formula.

```{r}
#| label: google-bootstrap-plot-render
#| renderings: [light, dark]
#| echo: false

google_bootstrap_plot_p + theme_narsil()
google_bootstrap_plot_p + theme_narsil_dark()
```

:::

::: {.fragment}

::: {.callout-tip appearance="simple"}
## When is bootstrap worth it?

Use bootstrap intervals when `gg_tsresiduals()` shows clearly non-normal residuals — heavy tails or marked skew. For series where residuals are roughly symmetric, the additional computational cost rarely changes the conclusion.
:::

:::

# Using STL Decomposition for forecasting with `decomposition_model()`

::: {.content-hidden unless-format="revealjs"}
## The idea
:::

::: {.content-visible unless-format="revealjs"}
### The idea
:::

STL (covered in [1.2](../02_ts_dcmp/ts_dcmp.qmd)) decomposes a series into trend-cycle, seasonal, and remainder components. The benchmarks (covered in [1.3](../03_fcst/forecasting.qmd)) each handle one aspect of the series well.

::: {.fragment}

The additive STL decomposition is

$$y_t = T_t + S_t + R_t$$

:::

::: {.fragment}

For forecasting we regroup it. Everything that is not seasonal goes into one term:

$$y_t = S_t + A_t, \qquad A_t = T_t + R_t$$

$A_t$ is the **seasonally adjusted** series. We now have two components, and we forecast each one with its own model:

$$\hat{y}_{T+h|T} = \hat{S}_{T+h|T} + \hat{A}_{T+h|T}$$

:::

::: {.fragment}

`decomposition_model()` turns this into a pipeline:

1. Apply STL to split $y_t$ into $S_t$ and $A_t$
2. Forecast the **seasonal component** $S_t$ — by default with SNAIVE, which just repeats the last seasonal cycle
3. Forecast the **seasonally adjusted series** $A_t$ with a model of your choice (here: Drift)
4. Add the two forecasts back together

:::

::: {.fragment}

| Component | Notation | `fable` column |
|---|---|---|
| Series | $y_t$ | the response variable (`Gas`) |
| Seasonal | $S_t$ | `season_year`, `season_week`, `season_day`, … — one per seasonal period detected by STL |
| Seasonally adjusted | $A_t$ | `season_adjust` |

The seasonal column is named after its period. Quarterly and monthly data give `season_year`; daily data can give `season_week` and `season_year` at the same time — we return to that in Module 4.

:::

::: {.fragment}

::: {.callout-note appearance="simple"}
The reconstruction is additive in the decomposition scale. If the STL was applied to a transformed series, the reconstruction happens before back-transformation.
:::

:::

::: {.content-hidden unless-format="revealjs"}

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-1
#| eval: false

stl_spec <- decomposition_model()
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-2
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season(), robust = TRUE),
)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-3
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season(), robust = TRUE),
  SNAIVE(season_year),
)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-4
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season(), robust = TRUE),
  SNAIVE(season_year),
  RW(season_adjust ~ drift())
)
```

## Syntax {auto-animate="true"}

```{r}
#| label: dcmp-animate-5
#| eval: false

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season(), robust = TRUE),
  SNAIVE(season_year),
  RW(season_adjust ~ drift())
)

gas_train |> model(stl_snaive_drift = stl_spec)
```

:::

::: {.content-visible unless-format="revealjs"}
### Syntax

The spec is defined once as a standalone object and passed into `model()`:
:::

::: {.content-hidden unless-format="revealjs"}
## Syntax {auto-animate="true"}
:::

```{r}
#| label: dcmp-fit

stl_spec <- decomposition_model(                  #<1>
  STL(Gas ~ trend() + season(), robust = TRUE),   #<2>
  SNAIVE(season_year),                            #<3>
  RW(season_adjust ~ drift())                     #<4>
)

gas_fit_dcmp <- gas_train |>
  model(stl_snaive_drift = stl_spec)              #<5>
```

1. `decomposition_model()` wraps a decomposition method and one or more component models into a single reusable spec.
2. STL decomposition — same syntax as in 1.2.
3. SNAIVE on `season_year` forecasts $S_t$. This line is optional — see the callout below.
4. Drift on `season_adjust` handles the seasonally adjusted series. `fable` reconstructs the final forecast by adding the two component forecasts.
5. We still need to put the decomposition model inside a `mable` with the `model()` function.

::: {.content-hidden unless-format="revealjs"}
## What if I omit the seasonal model?
:::

::: {.callout-note collapse="true"}
## Do I have to specify the seasonal model?

No. If you leave out the model for a seasonal component, `decomposition_model()` forecasts it with `SNAIVE()` automatically. The documentation puts it this way: all non-seasonal components must be specified, and any unspecified seasonal components are forecast using seasonal naive. Writing `SNAIVE(season_year)` explicitly, as we do here, changes nothing in the result — it just makes visible that the seasonal part is being modelled too, and shows where you would plug in a different model for it.

```{r}
#| label: dcmp-no-snaive
#| eval: false

decomposition_model(STL(Gas ~ trend() + season(), robust = TRUE), RW(season_adjust ~ drift()))
```
:::

::: {.content-hidden unless-format="revealjs"}

## With a transformation {auto-animate="true"}

```{r}
#| label: dcmp-animate-bc-1
#| eval: false
#| code-line-numbers: "2"

stl_spec <- decomposition_model(
  STL(Gas ~ trend() + season(), robust = TRUE),
  SNAIVE(season_year),
  RW(season_adjust ~ drift())
)

gas_train |> model(stl_snaive_drift = stl_spec)
```

:::

::: {.content-visible unless-format="revealjs"}
### With a transformation
:::

Box-Cox goes inside the STL spec — declared once, inherited everywhere:

```{r}
#| label: dcmp-fit-bc
#| code-line-numbers: "2"

stl_spec_bc <- decomposition_model(
  STL(box_cox(Gas, lambda) ~ trend() + season(), robust = TRUE), #<1>
  SNAIVE(season_year),
  RW(season_adjust ~ drift())
)

gas_fit_dcmp_bc <- gas_train |>
  model(stl_bc = stl_spec_bc)
```

1. The only difference from `stl_spec` is `box_cox()` in the STL call. All component models and the final back-transformation inherit it automatically.

```{r}
#| label: dcmp-fc-both
#| code-fold: true

gas_fc_dcmp <- bind_rows(                          #<1>
  gas_fit_dcmp    |> forecast(h = nrow(gas_test)),
  gas_fit_dcmp_bc |> forecast(h = nrow(gas_test))  #<2>
)
```

1. The two decomposition models live in separate mables, so forecast each and stack the fables.
2. Same horizon as every other model in this document, so the accuracy tables are comparable.

::: {.content-hidden unless-format="revealjs"}
## Final comparison
:::

::: {.content-visible unless-format="revealjs"}
### Final comparison
:::

All models side by side:

```{r}
#| label: gas-final-comparison
#| echo: false
#| output: false

gas_final_comparison_p <- bind_rows(
  gas_fc      |> filter(.model %in% c("snaive", "drift")),
  gas_fc_bc,
  gas_fc_dcmp
) |>
  autoplot(
    aus_production |> filter_index("2000 Q1" ~ .),
    level = NULL
  ) +
  labs(
    title = "Gas — all Module 1 models",
    y = "Petajoules", x = NULL
  )
```

```{r}
#| label: gas-final-comparison-render
#| renderings: [light, dark]
#| echo: false

plotly::ggplotly(gas_final_comparison_p + theme_narsil(), dynamicTicks = TRUE, width = 800, height = 400) |> gg_to_plotly_narsil()
plotly::ggplotly(gas_final_comparison_p + theme_narsil_dark(), dynamicTicks = TRUE, width = 800, height = 400) |> gg_to_plotly_narsil(dark = TRUE)
```

::: {.content-hidden unless-format="revealjs"}
## Accuracy table
:::

::: {.content-visible unless-format="revealjs"}
### Accuracy table
:::

```{r}
#| label: gas-final-accuracy
#| code-fold: true

gas_accu <- bind_rows(
  gas_fc      |> filter(.model %in% c("snaive", "drift")) |> accuracy(aus_production),
  gas_fc_bc   |> accuracy(aus_production),
  gas_fc_dcmp |> accuracy(aus_production)      #<1>
) |>
  select(.model, RMSE, MAE, MAPE, MASE, RMSSE) |>
  arrange(RMSSE)                               #<2>

gas_accu
```

1. All six Module 1 models, evaluated on the same test set with the full series passed to `accuracy()`.
2. Sorted by RMSSE: the first row is the baseline every later model has to beat.

::: {.fragment}

::: {.callout-important appearance="simple"}
## The baseline model

The best-performing model from this table is your benchmark for the rest of the course. Every model we build in Modules 2, 3, and 4 must beat this number to justify its additional complexity.
:::

:::

::: {.content-hidden unless-format="revealjs"}
## Refit on all data
:::

::: {.content-visible unless-format="revealjs"}
### Refit on all data
:::

Because the spec is a standalone object, refitting on the full dataset reuses it directly — no duplication:

```{r}
#| label: gas-final-refit
#| code-line-numbers: "1|4"
#| output: false

gas_final_fit <- aus_production |>
  model(stl_bc = stl_spec_bc)   #<1>

gas_final_fc <- gas_final_fit |>
  forecast(h = "2 years")

gas_final_refit_p <- gas_final_fc |>
  autoplot(aus_production |> filter_index("2000 Q1" ~ .)) +
  labs(
    title = "Australian gas production — STL + Box-Cox forecast",
    subtitle = "Refit on full data · 2-year horizon",
    y = "Petajoules", x = NULL
  )
```

```{r}
#| label: gas-final-refit-render
#| renderings: [light, dark]
#| echo: false

gas_final_refit_p + theme_narsil()
gas_final_refit_p + theme_narsil_dark()
```

1. Same spec, full data — no need to rewrite the model definition.

::: {.fragment}

::: {.callout-note appearance="simple"}
## Module 1 complete

You now have a principled, validated baseline model. It handles trend, seasonality, and non-constant variance. Modules 2 onward will replace or enhance individual components — but the workflow stays the same: split, fit, diagnose, compare, refit.
:::

:::
