---
title: "R, tsibble, and Time Series"
format:
  html: default
  revealjs:
    output-file: r_time_series_pres.html
---

# Packages

## The `tidyverse`

```{r}
#| label: pkg_tidyverse
#| message: true

library(tidyverse)

source(here::here("R/narsil_theme.R"))
theme_set(theme_narsil())
```

- `tidyverse` is a meta-package that loads the core packages of the [tidyverse](https://tidyverse.org/).

:::{.notes}

We will always load all the required packages a the beginning of the document. When loading the `tidyverse`, it shows which packages are being attached, as well as any conflicts with previously loaded packages.

:::{.callout-note collapse="true"}
## Core packages

- [`dplyr`](https://dplyr.tidyverse.org/) is the core package for **data transformation**. It is paired up with the following packages for specific column types:
  - [`stringr`](https://stringr.tidyverse.org/) for strings.
  - [`forcats`](https://forcats.tidyverse.org/) for **factors** (R's categorical data type).
  - [`lubridate`](https://lubridate.tidyverse.org/) for dates and date-times.
- [`ggplot2`](https://ggplot2.tidyverse.org/) is the primary package for visualization.
- [`readr`](https://readr.tidyverse.org/) is used to **import data** from delimited files (CSV, TSV, ...).
- [`tibble`](https://tibble.tidyverse.org/) is a modern reimagining of the data frame, keeping what time has proven to be effective, and throwing out what is not.
- [`tidyr`](https://tidyr.tidyverse.org/) is used to **tidy** data, i.e. to ensure that each variable is in its own column, each observation is in its own row, and each value is in its own cell.
- [`purrr`](https://purrr.tidyverse.org/) is used for functional programming with R.

:::

:::

## The `tidyverts`

```{r}
#| label: pkg_fpp3
#| message: true

library(fpp3)
```

- `fpp3` is also a meta-package that load the [tidyverts](https://tidyverts.org/) ecosystem for time series analysis and forecasting. 

:::{.notes}

The `tidyverts` packages are made to work seamlessly with the `tidyverse`.

:::{.callout-note collapse="true"}
 
- [`tsibble`](https://tsibble.tidyverts.org/) is the main data structure we will use to analyze and model time series. It is a **t**ime **series** t**ibble**.
- [`feasts`](https://feasts.tidyverts.org/) provides many functions and tools for feature and statistics extraction for time series.
- [`fable`](https://fable.tidyverts.org/) is the core package for modeling and foreasting time series.

:::

:::

# Time Series

## `tsibble` objects

A `tsibble` is a [`tibble`](https://tibble.tidyverse.org/index.html) that also records which column measures time and which columns identify each series.

Let's take a look at tourism in Australia:

```{r}
#| label: tourism_tsibble
#| paged-print: false
tourism
```

:::{.notes}

The `tsibble` has `{r} nrow(tourism)` rows and `{r} ncol(tourism)` columns. It shows quarterly data^[shown besides the tsibble dimension as `[1Q]`] on tourism across Australia.

Because the time structure belongs to the object rather than to our code, the `tidyverts` functions never have to guess where time lives: `autoplot()`, `model()`, and `forecast()` all read it from the tsibble. It also makes the object checkable — a tsibble refuses to be created if the same series has two observations at the same time point.

:::

## Index and key

```{r}
#| label: tourism_keys
#| paged-print: false

index_var(tourism) # <1>
key_vars(tourism)  # <2>
key_data(tourism)  # <3>
```

1. The **index** is the column along which time moves forward.
2. The **key** is the set of columns that distinguish one series from another. A tsibble can have no key at all (a single series), one key column, or several.
3. `key_data()` lists every distinct combination of the key columns, along with the rows belonging to each.

:::{.notes}

Index and key together define what one observation is: one point on the index, for one combination of the key. That pair is what a tsibble keeps unique, and those two arguments are all you need in order to build one.

`tourism` is divided by Region, State, and purpose of the trip^[these are specified in the `key` argument, and each distinct combination of them is one series. This tsibble contains `{r} n_keys(tourism)` of them.]. How many different states are there?

:::

## Australian States

```{r}
#| label: distinct_states

distinct(tourism, State)
```

## Which regions are located in Tasmania?

```{r}
#| label: distinct_regions

distinct(filter(tourism, State == "Tasmania"),Region)
```

## Building a tsibble

Most data does not arrive as a tsibble. `as_tsibble()` builds one from a data frame by naming the index and the key:

```{r}
#| label: sales_raw
#| paged-print: false

sales <- tibble(
  month = yearmonth(c("2024 Jan", "2024 Feb", "2024 Mar",   # <1>
                      "2024 Jan", "2024 Feb", "2024 Apr")),
  store = c("North", "North", "North", "South", "South", "South"),
  units = c(120, 135, 128, 98, 105, 111)
)

sales
```

1. `yearmonth()` turns text into a monthly time class. The index has to be a time class; a plain number or a string will not do.

```{r}
#| label: sales_as_tsibble
#| paged-print: false

sales_ts <- sales |>
  as_tsibble(index = month, key = store) # <1>

sales_ts
```

1. `month` is the index and `store` is the key. The header now reports `[1M]` for the monthly interval and `store [2]` for the two series it found.

:::{.notes}

`as_tsibble()` declares structure rather than changing values: the numbers are untouched, but the object now knows what time is. It also checks the declaration and fails immediately if a key–index pair is duplicated or if the index cannot be ordered. Getting that error at construction time is far cheaper than discovering the problem after fitting a model.

The interval `[1M]` is inferred, not declared. `tsibble` looks at the distances between consecutive index values and reports the finest regular interval that fits them.

:::

## Gaps and missing values

```{r}
#| label: sales_has_gaps
#| paged-print: false

has_gaps(sales_ts) # <1>
```

1. `South` jumps from February to April, so March is absent as a *row* rather than present with a missing value.

```{r}
#| label: sales_fill_gaps
#| paged-print: false

sales_ts |>
  fill_gaps() # <1>
```

1. `fill_gaps()` inserts the missing row and marks `units` as `NA`, turning the implicit gap into an explicit missing value.

:::{.notes}

A missing value is a row that exists with `NA` in it. A gap is a row that does not exist at all. The distinction matters because a series with an invisible hole in it will be read as if February were followed by April with nothing in between, so lags, seasonal periods, and differences all shift by one position from that point onward.

`fill_gaps()` is what makes the problem visible. Once the gap is an `NA`, you can decide what to do about it: interpolate it, drop the affected span, or leave it and let the model handle it.

:::

## Data Transformation: Average trips

To get the average trips by purpose, we need to do the following:

:::{.incremental}

1) Filter the original `tsibble` to get only the data from East Coast, Tasmania.
1) Convert the data to a `tibble`.
1) Group by purpose.
1) Summarise by getting the mean of the trips.

:::

## {auto-animate="true"}

With traditional code, this would look something like:

```{r}
#| label: no_pipe_summarise
#| eval: false

summarise(group_by(as_tibble(filter(tourism, State == "Tasmania", 
                                    Region == "East Coast")), Purpose),
          mean_trips = mean(Trips))
```

:::{.notes}
:::{.callout-note collapse="true"}
## Order of code execution
Note that this code must be read inside-out. This makes it harder to understand, and also harder to debug.
:::
:::

:::{.content-hidden unless-format="revealjs"}

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_1
#| eval: false

tourism |>                          
  filter() 
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_2
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_3
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
  as_tibble() |>
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_4
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
  as_tibble() |>                    
  group_by() 
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_5
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
  as_tibble() |>                    
  group_by(Purpose) |>
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_6
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
  as_tibble() |>                    
  group_by(Purpose) |>              
  summarise(mean_trips = )
```

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_7
#| eval: false

tourism |>                          
  filter(State == "Tasmania",       
         Region == "East Coast") |> 
  as_tibble() |>                    
  group_by(Purpose) |>              
  summarise(mean_trips = mean(Trips))
```

:::

## {auto-animate="true"}

Using the native pipe operator; ` |> `, we can improve the same code:

```{r}
#| label: pipe_summarise_full

tourism |>                            # <1>
  filter(State == "Tasmania",         # <2>
         Region == "East Coast") |>   # <2>
  as_tibble() |>                      # <3>
  group_by(Purpose) |>                # <4>
  summarise(mean_trips = mean(Trips)) # <5>
```

1. Take the tsibble `tourism`, *then*
2. filter by State and Region, *then*
3. convert to a `tibble`, *then*
4. group the tibble by purpose, *then*
5. summarise by taking the mean trips

:::{.notes}
:::{.callout-tip collapse="true"}
## The pipe operator |> 

The pipe is read as "**then**", and it allows us to write code in the order it's supposed to be run.

It also helps to debug code easier, because you can run each function in order and see where the error is.
:::
:::
# TS Visualization

:::{.content-hidden unless-format="revealjs"}

## Plotting tourism across time {auto-animate="true"}

```{r}
#| label: ts_viz_1
#| eval: true

tourism
```

## Plotting tourism across time {auto-animate="true"}

```{r}
#| label: ts_viz_2
#| eval: true
#| code-line-numbers: "2-3"

tourism |> 
  filter(State == "Tasmania",
         Region == "East Coast")               
```

## Plotting tourism across time {auto-animate="true"}

```{r}
#| label: ts_viz_3
#| eval: true
#| code-line-numbers: "4"

tourism |> 
  filter(State == "Tasmania",
         Region == "East Coast") |> 
  autoplot(Trips)                
```

## Plotting tourism across time {auto-animate="true"}

```{r}
#| label: ts_viz_4
#| eval: true
#| code-line-numbers: "5"

tourism |> 
  filter(State == "Tasmania",
         Region == "East Coast") |> 
  autoplot(Trips) +
  facet_wrap(vars(Purpose), scale = "free_y")                
```

:::

## Plotting tourism across time {auto-animate="true"}

```{r}
#| label: ts_viz_full
#| output: false

tourism_p <- tourism |>
  filter(State == "Tasmania",
         Region == "East Coast") |>
  autoplot(Trips) +                             # <1>
  facet_wrap(vars(Purpose), scale = "free_y") + # <2>
  theme(legend.position = "none")               # <3>
```

1. `autoplot()` detects the data automatically and proposes a plot accordingly.
2. `facet_wrap()` Divides a plot into subplots (facets).
3. you can customize endless feautres using `theme()`. Here, we remove the legend, as it's redudant.

```{r}
#| label: ts_viz_full_render
#| renderings: [light, dark]
#| echo: false

tourism_p + scale_narsil() + theme(legend.position = "none")
tourism_p + scale_narsil_dark() + theme(legend.position = "none")
```

## Time plots

:::{.columns}

::: {.column width="50%"}

```{r}
#| label: time_plot_1
#| code-line-numbers: "2"
#| output: false

gas_p <- aus_production |>
  autoplot(Gas)
```

```{r}
#| label: time_plot_1_render
#| renderings: [light, dark]
#| echo: false

gas_p + scale_narsil()
gas_p + scale_narsil_dark()
```

:::

::: {.column width="50%"}

```{r}
#| label: time_plot_2
#| code-line-numbers: "2-3"
#| output: false

gas_points_p <- aus_production |>
  autoplot(Gas) +
  geom_point()
```

```{r}
#| label: time_plot_2_render
#| renderings: [light, dark]
#| echo: false

gas_points_p + scale_narsil()
gas_points_p + scale_narsil_dark()
```

:::
:::

:::{.notes}

These are the most basic type of plots. We have the time variable in the $x$-axis, and our forecast variable in the $y$-axis. Time plots should be line plots, and can include or not points.

:::

## Seasonal Plots

```{r}
#| label: gg_season
#| code-line-numbers: "2"
#| output: false

gas_season_p <- aus_production |>
  gg_season(Gas)
```

```{r}
#| label: gg_season_render
#| renderings: [light, dark]
#| echo: false

gas_season_p + theme_narsil()
gas_season_p + theme_narsil_dark()
```

:::{.notes}

The data here are plotted against a single "season". It's useful in identifying years with changes in patterns. 

:::

Removing the trend from the data:

```{r}
#| label: gg_season2
#| echo: false
#| output: false

gas_sin_tend_p <- aus_production |>
  model(STL(Gas, robust = TRUE)) |>
  components() |>
  mutate(sin_tend = season_year + remainder) |>
  gg_season(sin_tend)
```

```{r}
#| label: gg_season2_render
#| renderings: [light, dark]
#| echo: false

gas_sin_tend_p + theme_narsil()
gas_sin_tend_p + theme_narsil_dark()
```


## Seasonal Subseries Plots

```{r}
#| label: gg_subseries
#| code-line-numbers: "2"
#| output: false

gas_subseries_p <- aus_production |>
  gg_subseries(Gas)
```

```{r}
#| label: gg_subseries_render
#| renderings: [light, dark]
#| echo: false

gas_subseries_p + scale_narsil()
gas_subseries_p + scale_narsil_dark()
```

:::{.notes}

Here we split the plot into many subplots, one for each season. This helps us see clearly the underlying seasonal pattern. The mean for each season is represented as the blue horizontal line.

:::

## `gg_tsdisplay()`

```{r}
#| label: gg_tsdisplay
#| code-line-numbers: "2"
#| results: hide

theme_narsil()
aus_production |>
  gg_tsdisplay(Gas, plot_type = "season")
```

:::{.notes}

This function provides a convenient way to have 3 plots: a time plot, an ACF plot, and a third option that can be customized with one of the following plot types:

:::{.callout-tip collapse="true"}
## `plot_type` options

- "auto", 
- "partial", 
- "season", 
- "histogram", 
- "scatter", 
- "spectrum"

:::
:::

## Exporting data to .csv

```{r}
#| label: write_csv
#| eval: false
#| code-line-numbers: "5"

tourism |> 
  filter(State == "Tasmania",
         Region == "East Coast") |> 
  mutate(Quarter = as.Date(Quarter)) |> 
  write_csv("./datos/tasmania.csv")
```

:::{.notes}
You can export to .csv by providing a `tsibble` or `tibble` (or any other type of data frame), by calling **`write_csv()`**, and specifying the output file's name.
:::

