---
title: "Preprocessing Regressors for Forecasting"
description: "A practical guide to preparing explanatory variables before fitting a TSLM: scaling, transformations, categorical encoding, interactions, and a pre-flight checklist."
date: 2026-03-28
categories:
  - stats
---
```{r}
#| label: setup
#| message: false
#| include: false

library(tidyverse)
library(fpp3)
```

This document is a companion to the

[Linear Regression Models](../../modules/module_3/02_regression/regression.qmd)
module. It addresses questions that frequently come up when building a `TSLM()`:

*Do I need to scale my variables? Should I log-transform this regressor?
How do I handle a categorical variable with more than two levels?*

The short answer to all three: *it depends* — and this document explains on what.

# Scaling and Standardization

## What scaling does (and does not do)

Standardization transforms a variable $x$ into:

$$
z = \frac{x - \bar{x}}{s_x}
$$

so that $z$ has mean 0 and standard deviation 1. Min-max normalization maps
the variable to $[0, 1]$:

$$
z = \frac{x - x_{\min}}{x_{\max} - x_{\min}}
$$

Both are linear transformations. Because OLS estimates are equivariant under
linear transformations of the regressors, **scaling does not change predicted
values, residuals, or goodness-of-fit metrics**. It only rescales the
coefficients.

::: {.callout-important}
## Key takeaway

Scaling is **irrelevant for forecast quality** in OLS-based models like
`TSLM()`. Predicted values $\hat{y}$ are identical whether you scale or not.
:::

## When scaling *does* matter

There are two situations where scaling genuinely changes the result:

**Regularized regression (Ridge, Lasso, Elastic Net).** The penalty term
$\lambda \sum_j \beta_j^2$ treats all coefficients equally. If $x_1$ is
measured in millions of pesos and $x_2$ is a rate between 0 and 1, the
penalty will shrink $\hat\beta_2$ much more aggressively than $\hat\beta_1$
simply because of scale differences — not because $x_2$ is less important.
Scaling before regularization is therefore **required**.

**Comparing coefficient magnitudes.** If you want to interpret which regressor
has the largest *effect size* (not just the largest coefficient), standardized
coefficients (also called beta coefficients) make the comparison meaningful.
This is an inferential use, not a forecasting use.

::: {.callout-note}
## `TSLM()` does not regularize

`fable`'s `TSLM()` uses ordinary least squares without any penalty term.
Scaling is therefore unnecessary for `TSLM()` models.
:::

## A quick illustration
```{r}
#| label: scaling-demo
#| message: false

us_change_scaled <- us_change |>                                # <1>
  mutate(across(                                                # <2>
    c(Income, Production, Savings, Unemployment),
    \(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE),   # <3>
    .names = "{.col}_z"                                        # <3>
  ))

fit_original <- us_change |>                                   # <4>
  model(TSLM(Consumption ~ Income + Production))

fit_scaled <- us_change_scaled |>                              # <5>
  model(TSLM(Consumption ~ Income_z + Production_z))
```

1. Start from `us_change`, the dataset used throughout the regression module.
2. `across()` applies the same transformation to multiple columns at once.
3. The anonymous function `\(x)` standardizes each selected column. `.names = "{.col}_z"` creates new columns with a `_z` suffix, preserving the originals.
4. Fit the model on the original scale.
5. Fit the same model on the standardized scale.

```{r}
#| label: scaling-forecast-comparison
#| message: false

fc_original <- fit_original |>                                 # <1>
  forecast(us_change |> tail(8))

fc_scaled <- fit_scaled |>                                     # <2>
  forecast(us_change_scaled |> tail(8))

bind_cols(                                                     # <3>
  fc_original |> as_tibble() |> select(.mean) |> rename(original = .mean),
  fc_scaled   |> as_tibble() |> select(.mean) |> rename(scaled   = .mean)
)
```

1. Forecast using the original-scale model.
2. Forecast using the standardized model.
3. Side-by-side comparison. The columns should be identical (within floating-
   point precision).

::: {.callout-tip}
## Try it yourself

Compute `accuracy()` for both models. Do the RMSE and MAE differ?
:::

# Transforming the Response and Regressors

## Transforming $y$: back-transformation and bias

You already saw Box-Cox and log transformations applied to $y$ in the
[decomposition module](../../modules/module_1/02_ts_dcmp/ts_dcmp.qmd).
The same logic applies inside a regression: if your series has
multiplicative seasonality or variance that grows with the level, transforming
$y$ before fitting can stabilize residuals and improve forecast intervals.

In `fable`, the transformation is applied directly inside the model formula
and back-transformation is handled automatically:
```{r}
#| label: transform-y
#| eval: false

model(TSLM(log(Consumption) ~ Income + Production))
```

::: {.callout-warning}
## Back-transformation introduces bias

When you back-transform a forecast from log scale, $e^{\hat{y}}$ is the
median of the forecast distribution, not the mean. `fable` corrects for this
automatically when using `forecast()`, but it is worth knowing when
interpreting point forecasts from other tools.
:::

## Transforming regressors: linearizing relationships

Transforming a *regressor* $x$ is about **linearizing the relationship**
between $x$ and $y$. OLS assumes the relationship is linear in the
parameters — but the regressors themselves can be nonlinear functions of
the original variables.

Common cases:

| Relationship | Transform | Formula |
|---|---|---|
| Diminishing returns | Log | $y \sim \log(x)$ |
| Multiplicative | Both log | $\log(y) \sim \log(x)$ |
| Quadratic | Polynomial | $y \sim x + x^2$ |
| Exponential growth | Log $y$ only | $\log(y) \sim x$ |
```{r}
#| label: transform-regressor
#| eval: false

model(TSLM(Consumption ~ log(Income) + Production))            # <1>
model(TSLM(Consumption ~ Income + I(Income^2)))                # <2>
```

1. Log-transform `Income` to capture diminishing marginal effects.
2. Add a quadratic term for `Income`. The `I()` wrapper tells R to treat
   `Income^2` arithmetically, not as a formula interaction.

::: {.callout-note}
## This is different from transforming $y$

Transforming a regressor changes the *shape* of the relationship between
$x$ and $y$. Transforming $y$ changes the *distributional assumptions*
about the residuals. Both can be done simultaneously.
:::

# Categorical Variables

## From categories to numbers: dummy encoding

OLS cannot directly use a text variable. A categorical variable with $k$
levels is converted into $k - 1$ binary (0/1) dummy variables. The omitted
level becomes the **reference category**, and every coefficient is interpreted
relative to it.

::: {.callout-important}
## Why $k - 1$ and not $k$?

Including all $k$ dummies would create perfect multicollinearity (the
*dummy variable trap*): the $k$ columns always sum to 1, which is identical
to the intercept column. OLS cannot invert the design matrix in that case.
:::

In R, factors are automatically expanded into dummies by `lm()` and
therefore by `TSLM()`. You rarely need to create them manually:
```{r}
#| label: dummy-auto
#| message: false
#| eval: false

us_employment |>
  filter(Title %in% c("Mining and Logging", "Construction", "Manufacturing")) |>
  model(TSLM(Employed ~ trend() + season() + Title))           # <1>
```

1. `Title` is a character/factor variable. `TSLM()` will automatically
   create two dummies (three levels minus one reference).

## Choosing the reference category

R picks the reference level alphabetically by default. You can change it
with `relevel()` if a different baseline makes more interpretive sense:
```{r}
#| label: relevel
#| eval: false

data |>
  mutate(quarter = relevel(factor(quarter), ref = "Q1")) |>    # <1>
  model(TSLM(y ~ trend() + quarter))
```

1. Set Q1 as the reference so that all quarterly dummies are interpreted
   as deviations from Q1.

::: {.callout-tip}
## `season()` in `TSLM()` already handles this

When you use `season()` inside `TSLM()`, `fable` creates the seasonal
dummies for you and drops one level automatically. You only need manual
dummy encoding when your categorical variable is *not* the time index's
seasonal period — for example, a product category, a region, or a
promotional event type.
:::

# Interactions and Nonlinear Terms

## What an interaction says

An interaction $x_1 \times x_2$ allows the *effect* of $x_1$ on $y$ to
depend on the value of $x_2$. Without an interaction, OLS assumes the two
effects are additive and independent.

In forecasting terms: maybe the effect of a promotional discount (`promo`)
on sales is larger during the holiday season (`holiday = 1`) than during
the rest of the year. That is an interaction.
```{r}
#| label: interaction
#| eval: false

model(TSLM(Sales ~ trend() + season() + promo * holiday))      # <1>
model(TSLM(Sales ~ trend() + season() + promo + holiday +
             promo:holiday))                                   # <2>
```

1. The `*` shorthand expands to main effects plus the interaction term.
2. The `:` operator adds only the interaction term, assuming the main
   effects are already included.

::: {.callout-warning}
## Interactions increase model complexity fast

With $p$ regressors, you can form $\binom{p}{2}$ pairwise interactions.
Adding all of them will almost certainly overfit, especially with short
time series. Only include interactions that have a clear theoretical
justification.
:::

## Polynomial terms

For a smooth nonlinear relationship, polynomial regression adds powers of
a regressor:

$$
y_t = \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \varepsilon_t
$$
```{r}
#| label: polynomial
#| eval: false

model(TSLM(y ~ x + I(x^2) + I(x^3)))                         # <1>
```

1. Always wrap polynomial terms in `I()` inside a formula. Without it,
   `^` has a different meaning in R's formula syntax.

::: {.callout-note}
## Polynomial vs. splines

High-degree polynomials oscillate wildly near the boundaries of the data.
Splines (piecewise polynomials with smooth joins) are generally preferred
for capturing smooth nonlinearities — see `splines::bs()` or `splines::ns()`
for use inside `lm()` and `TSLM()`.
:::

# Pre-flight Checklist

Before running a `TSLM()`, work through these questions:

| Question | Action |
|---|---|
| Does $y$ have variance that grows with its level? | Apply Box-Cox or log to $y$ |
| Does a regressor have a nonlinear relationship with $y$? | Transform or add polynomial/spline terms |
| Does a regressor have very different scale from the others? | No action needed for OLS; scale only if using regularization |
| Do you have a categorical regressor? | Let `TSLM()` handle it automatically (factor/character columns) |
| Does the effect of one regressor depend on another? | Add an interaction term — only if theoretically justified |
| Do you have many candidate regressors? | Check VIF for multicollinearity; consider regularization |

::: {.callout-important}
## The most common mistake

Running a TSLM with standardized variables because "it's good practice" —
and then being confused by uninterpretable coefficients. Scale only when
you have a specific reason to do so.
:::

::: {.callout-tip collapse="true"}
## Quick reference: R formula syntax

| Syntax | Meaning |
|---|---|
| `y ~ x1 + x2` | Main effects only |
| `y ~ x1 * x2` | Main effects + interaction |
| `y ~ x1:x2` | Interaction only |
| `y ~ I(x^2)` | Arithmetic expression (polynomial) |
| `y ~ log(x)` | Function of a regressor |
| `y ~ .` | All remaining columns as regressors |
:::
