---
title: "Regression models"
format:
  html: default
  revealjs:
    output-file: reg_pres.html
draft: true
---

```{r}
#| label: pkgs
#| message: false

library(tidyverse)
library(fpp3)
library(ggtime)
```


## Regression Models for Time Series forecasting

::: {.notes}

### Introduction

Regression models can be powerful tools for understanding and forecasting time series when the relationship between a dependent variable and one or more explanatory variables is approximately linear. Unlike ARIMA models, regression models explicitly model deterministic components such as trend and seasonality.

:::

::: {.content-hidden unless-format="html"}

### Introduction

Regression models can help identify and forecast deterministic patterns in time series. They are particularly useful when explanatory variables or known calendar effects drive much of the observed variation.

:::

---

### Time Series Regression Models

A **time series regression model** expresses a variable $y_t$ as a linear combination of explanatory variables:

$$
y_t = \beta_0 + \beta_1 x_{1t} + \beta_2 x_{2t} + \dots + \varepsilon_t
$$

where $\varepsilon_t$ is the random error term.

We will use the `us_change` dataset from the *fpp3* package, which contains quarterly percentage changes in US consumption, income, production, savings, and unemployment.

```{r}
library(fpp3)

us_change |> 
  glimpse()
```

A simple regression model relating Consumption to Income can be fit using:

```{r}
us_change |>
model(OLS = TSLM(Consumption ~ Income)) |>
report()
```
::: {.callout-tip collapse="true"}
## Interpretation

The estimated coefficient for Income represents the average change in consumption (percentage points) associated with a one–percentage-point change in income.

:::

