---
title: "Modular Forecasting: Mixing ETS and ARIMA"
format:
  html: default
  revealjs:
    output-file: module_2_synthesis_pres.html
freeze: false
---

```{r}
#| label: pkgs
#| message: false
#| include: false

library(tidyverse)
library(fpp3)
```

:::{.content-visible unless-format="revealjs"}

```{r}
#| label: pkgs_special
#| message: false

library(tidyquant) #<1>
library(plotly)    #<1>
```

1. In addition to the regular packages, here we'll use `tidyquant` and `plotly`.

:::

:::{.callout-note collapse="true"}
## Data setup: `mexretail`

`mexretail` was first introduced in [Module 1 — Time Series Decomposition](../../module_1/02_ts_dcmp/ts_dcmp.qmd#ts-features--patterns) and has been our running dataset throughout Module 2.

```{r}
#| label: data-setup
#| message: false

mexretail <- tq_get(
  "MEXSLRTTO01IXOBM",
  get = "economic.data",
  from = "1985-01-01"
) |>
  mutate(date = yearmonth(date)) |>
  rename(y = price) |>
  as_tsibble(index = date)

mexretail_train <- mexretail |> filter(year(date) <= 2019)
mexretail_test  <- mexretail |> filter(year(date) >  2019)
```

:::

# Where We Are

:::{.content-hidden unless-format="revealjs"}
##
:::

Module 2 has added two new model families to our toolkit. Together with `decomposition_model()`, they give us far more combinations than it might initially seem:

:::{.content-visible unless-format="revealjs"}

| Approach | Decomposition | Trend-cycle model | Seasonal model |
|:--|:--:|:--:|:--:|
| **M1 baseline** | STL | Drift | SNAIVE |
| **STL + ETS** | STL | ETS | SNAIVE |
| **STL + ARIMA** | STL | ARIMA | SNAIVE |
| **STL + ETS + ARIMA** | STL | ETS | ARIMA |
| **STL + ARIMA + ETS** | STL | ARIMA | ETS |
| **ETS** | — | ETS | ETS |
| **SARIMA** | — | ARIMA | ARIMA$(P,D,Q)_m$ |

:::

:::{.content-hidden unless-format="revealjs"}

```{r}
#| label: module-map-table
#| echo: false

tibble::tribble(
  ~Approach,               ~Decomposition, ~`Trend-cycle`,  ~`Seasonal`,
  "M1 baseline",           "STL",          "Drift",         "SNAIVE",
  "STL + ETS",             "STL",          "ETS",           "SNAIVE",
  "STL + ARIMA",           "STL",          "ARIMA",         "SNAIVE",
  "STL + ETS + ARIMA",     "STL",          "ETS",           "ARIMA",
  "STL + ARIMA + ETS",     "STL",          "ARIMA",         "ETS",
  "ETS",                   "—",            "ETS",           "ETS",
  "SARIMA",                "—",            "ARIMA",         "ARIMA(P,D,Q)m"
) |>
  knitr::kable()
```

:::

:::{.fragment}
:::{.callout-note appearance="simple"}
## `decomposition_model()` is more flexible than it looked

In the previous sections, we always let STL handle the seasonal component implicitly — which defaults to SNAIVE. But `decomposition_model()` lets you specify a model for *each* component separately. This opens up the mixed combinations in the middle of the table, and that is exactly what we explore here.
:::
:::

# Mixing ETS and ARIMA inside `decomposition_model()`

:::{.content-hidden unless-format="revealjs"}
## The structure
:::

Recall the structure of `decomposition_model()`:

```r
decomposition_model(
  <decomposition method>,
  <model for season_adjust>,   # trend-cycle component
  <model for season_year>      # seasonal component (defaults to SNAIVE if omitted)
)
```

:::{.fragment}
Each component has its own characteristics, and we can choose the best model for each one *independently*:

:::{.callout-tip appearance="simple"}
## The constraint logic

| Component | What it contains | What to suppress |
|:--|:--|:--|
| `season_adjust` | trend + cycle + remainder | seasonality → `season("N")` / `PDQ(0,0,0)` |
| `season_year` | seasonal pattern only | trend → `trend("N")` |

:::
:::

:::{.content-hidden unless-format="revealjs"}
## Defining the model specs {auto-animate="true"}
:::

:::{.content-visible unless-format="revealjs"}
## Defining the model specs
:::

Rather than embedding the full `decomposition_model()` calls inside the `mable`, we define them as named objects first. This keeps the model comparison code clean and makes each specification easy to inspect and reuse.

```{r}
#| label: baseline-spec

baseline_spec <- decomposition_model(        #<1>
  STL(log(y) ~
        trend(window = NULL) +
        season(window = "periodic"),
      robust = TRUE),
  RW(season_adjust ~ drift()),
  SNAIVE(season_year)
)
```

1. Module 1 baseline: STL + Drift + SNAIVE. Our benchmark — every model must beat this. See [The Forecasting Workflow](../../forecasting_workflow.qmd).

```{r}
#| label: stl-ets-spec

stl_ets_spec <- decomposition_model(         #<1>
  STL(log(y) ~
        trend(window = NULL) +
        season(window = "periodic"),
      robust = TRUE),
  ETS(season_adjust ~ season("N"))           #<2>
)
```

1. STL + ETS on the seasonally adjusted series; seasonal component defaults to SNAIVE. See [Exponential Smoothing](../01_ets/ets.qmd).
2. `season("N")` suppresses seasonality — STL already removed it from `season_adjust`.

```{r}
#| label: stl-arima-spec

stl_arima_spec <- decomposition_model(       #<1>
  STL(log(y) ~
        trend(window = NULL) +
        season(window = "periodic"),
      robust = TRUE),
  ARIMA(season_adjust ~ PDQ(0, 0, 0))        #<2>
)
```

1. STL + ARIMA on the seasonally adjusted series; seasonal component defaults to SNAIVE. See [ARIMA Models](../03_arima/arima.qmd#stl--arima).
2. `PDQ(0,0,0)` suppresses seasonal ARIMA terms — not needed on the seasonally adjusted series.

:::{.content-hidden unless-format="revealjs"}
## STL + ETS + ARIMA {auto-animate="true"}
:::

:::{.content-visible unless-format="revealjs"}
## STL + ETS + ARIMA
:::

ETS models the trend-cycle; ARIMA models the seasonal pattern:

```{r}
#| label: stl-ets-arima-spec

stl_ets_arima_spec <- decomposition_model(
  STL(log(y) ~
        trend(window = NULL) +
        season(window = "periodic"),
      robust = TRUE),
  ETS(season_adjust ~ season("N")),          #<1>
  ARIMA(season_year)
)
```

1. ETS with `season("N")` on the trend-cycle — no seasonality to model here.

:::{.content-hidden unless-format="revealjs"}
## STL + ARIMA + ETS {auto-animate="true"}
:::

:::{.content-visible unless-format="revealjs"}
## STL + ARIMA + ETS
:::

ARIMA models the trend-cycle; ETS models the seasonal pattern:

```{r}
#| label: stl-arima-ets-spec

stl_arima_ets_spec <- decomposition_model(
  STL(log(y) ~
        trend(window = NULL) +
        season(window = "periodic"),
      robust = TRUE),
  ARIMA(season_adjust ~ PDQ(0, 0, 0)),       #<1>
  ETS(season_year ~ trend("N"))              #<2>
)
```

1. ARIMA with `PDQ(0,0,0)` on the trend-cycle — no seasonal ARIMA terms needed.
2. ETS with `trend("N")` on the seasonal component — the seasonal pattern has no trend.

# Revisiting `mexretail`

:::{.content-hidden unless-format="revealjs"}
## Fitting all approaches
:::

With the specs defined, the `mable` is straightforward to read — each row is a named approach:

```{r}
#| label: mexretail-fit
#| message: false
#| warning: false

mexretail_fit <- mexretail_train |>
  model(
    baseline      = baseline_spec,
    stl_ets       = stl_ets_spec,
    stl_arima     = stl_arima_spec,
    stl_ets_arima = stl_ets_arima_spec,
    stl_arima_ets = stl_arima_ets_spec,
    ets           = ETS(log(y)),                             #<1>
    sarima        = ARIMA(log(y),                            #<2>
                          stepwise      = FALSE,
                          approximation = FALSE)
  )

mexretail_fit
```

1. Full ETS without explicit decomposition — Holt-Winters handles everything internally. See [Exponential Smoothing](../01_ets/ets.qmd#methods-with-seasonality).
2. Full SARIMA without explicit decomposition. See [ARIMA Models](../03_arima/arima.qmd#sarima).

:::{.content-hidden unless-format="revealjs"}
## Forecasts
:::

```{r}
#| label: mexretail-fc
#| message: false
#| warning: false

mexretail_fc <- mexretail_fit |>
  forecast(h = nrow(mexretail_test))
```

:::{.content-visible unless-format="revealjs"}

```{r}
#| label: mexretail-fc-plot
#| echo: false
#| message: false
#| warning: false

p <- mexretail_fc |>
  autoplot(mexretail |> filter(year(date) >= 2015), level = 80) +
  facet_wrap(vars(.model), ncol = 2) +
  labs(
    title    = "mexretail — all approaches compared",
    y        = "Index 2015=100",
    subtitle = "80% prediction intervals"
  )

ggplotly(p, dynamicTicks = TRUE)
```

:::

:::{.content-hidden unless-format="revealjs"}

```{r}
#| label: mexretail-fc-plot-pres
#| echo: false
#| message: false
#| warning: false

mexretail_fc |>
  autoplot(mexretail |> filter(year(date) >= 2015), level = 80) +
  facet_wrap(vars(.model), ncol = 2) +
  labs(
    title    = "mexretail — all approaches compared",
    y        = "Index 2015=100",
    subtitle = "80% prediction intervals"
  )
```

:::

:::{.content-hidden unless-format="revealjs"}
## Accuracy on the test set
:::

```{r}
#| label: mexretail-accuracy
#| message: false
#| warning: false

mexretail_accu <- mexretail_fc |>
  accuracy(mexretail) |>
  select(.model, RMSE, MAE, MAPE, MASE) |>
  arrange(RMSE)

mexretail_accu
```

:::{.fragment}
:::{.callout-tip appearance="simple"}
## Reading the table

A MASE below 1 means the model beats the SNAIVE benchmark. Check where `baseline` lands — that is the floor set in Module 1. Everything above it in this ranking is a genuine improvement.
:::
:::

# A New Series: Australian Beer Production

:::{.content-hidden unless-format="revealjs"}
## Setup
:::

To check whether the rankings hold on a different series, we apply the same workflow to quarterly Australian beer production from `fpp3`:

```{r}
#| label: beer-setup
#| message: false

beer_train <- aus_production |>
  rename(y = Beer) |>                  #<1>
  select(Quarter, y) |>
  filter(!is.na(y), year(Quarter) <= 2006)

beer_test <- aus_production |>
  rename(y = Beer) |>
  select(Quarter, y) |>
  filter(!is.na(y), year(Quarter) > 2006)

beer_full <- bind_rows(beer_train, beer_test)
```

1. We rename `Beer` to `y` — the same convention used for `mexretail`. This is what allows the specs defined above to be reused directly.

```{r}
#| echo: false

beer_train |>
  autoplot(y, color = "goldenrod3") +
  labs(
    title = "Australian beer production (train set)",
    y     = "Megalitres",
    x     = ""
  )
```

:::{.callout-tip appearance="simple"}
## Reusing model specs across series

Because both `mexretail` and `beer` use `log(y)` as the response variable, the specs defined above apply to both series without any modification. The `mable` for `beer` will look identical to the one for `mexretail` — the only difference is the `tsibble` being piped in.

This is the payoff of two conventions working together: naming your response variable `y`, and defining specs outside the `mable`.
:::

:::{.content-hidden unless-format="revealjs"}
## Fitting all approaches
:::

```{r}
#| label: beer-fit
#| message: false
#| warning: false

beer_fit <- beer_train |>
  model(
    baseline      = baseline_spec,
    stl_ets       = stl_ets_spec,
    stl_arima     = stl_arima_spec,
    stl_ets_arima = stl_ets_arima_spec,
    stl_arima_ets = stl_arima_ets_spec,
    ets           = ETS(log(y)),
    sarima        = ARIMA(log(y), stepwise = FALSE, approximation = FALSE)
  )

beer_fit
```

:::{.content-hidden unless-format="revealjs"}
## Forecasts and accuracy
:::

```{r}
#| label: beer-fc
#| message: false
#| warning: false

beer_fc <- beer_fit |>
  forecast(h = nrow(beer_test))
```

:::{.content-visible unless-format="revealjs"}

```{r}
#| label: beer-fc-plot
#| echo: false
#| message: false
#| warning: false

p <- beer_fc |>
  autoplot(beer_full |> filter(year(Quarter) >= 2000), level = 80) +
  facet_wrap(vars(.model), ncol = 2) +
  labs(
    title    = "Australian beer production — all approaches compared",
    y        = "Megalitres",
    subtitle = "80% prediction intervals"
  )

ggplotly(p, dynamicTicks = TRUE)
```

:::

:::{.content-hidden unless-format="revealjs"}

```{r}
#| label: beer-fc-plot-pres
#| echo: false
#| message: false
#| warning: false

beer_fc |>
  autoplot(beer_full |> filter(year(Quarter) >= 2000), level = 80) +
  facet_wrap(vars(.model), ncol = 2) +
  labs(
    title    = "Australian beer — all approaches",
    y        = "Megalitres",
    subtitle = "80% prediction intervals"
  )
```

:::

```{r}
#| label: beer-accuracy
#| message: false
#| warning: false

beer_accu <- beer_fc |>
  accuracy(beer_full) |>
  select(.model, RMSE, MAE, MAPE, MASE) |>
  arrange(RMSE)

beer_accu
```

:::{.fragment}
:::{.callout-note appearance="simple"}
## Does the winner change?

Compare the ranking here against `mexretail`. Does the same approach win on both series? What does that tell you about when mixed decomposition models add value — and when they don't?
:::
:::

# When to Use Which

:::{.content-hidden unless-format="revealjs"}
##
:::

:::{.content-visible unless-format="revealjs"}

| Approach | Seasonal pattern | Outlier robustness | Best when... |
|:--|:--|:--|:--|
| **STL + ETS** | Flexible, can evolve | High | Trend is smooth; seasonality stable |
| **STL + ARIMA** | Flexible, can evolve | High | Trend has autocorrelation structure |
| **STL + ETS + ARIMA** | Flexible, can evolve | High | Trend smooth; seasonal pattern has autocorrelation |
| **STL + ARIMA + ETS** | Flexible, can evolve | High | Trend autocorrelated; seasonal is level-shifting |
| **ETS** | Fixed structure | Moderate | Short series; no need for explicit decomposition |
| **SARIMA** | Fixed structure | Lower | Compact model preferred; longer series |

:::

:::{.content-hidden unless-format="revealjs"}

```{r}
#| label: when-to-use-table
#| echo: false

tibble::tribble(
  ~Approach,              ~`Seasonal pattern`,        ~`Outlier robustness`, ~`Best when...`,
  "STL + ETS",            "Flexible, can evolve",     "High",                "Trend smooth; seasonality stable",
  "STL + ARIMA",          "Flexible, can evolve",     "High",                "Trend has autocorrelation structure",
  "STL + ETS + ARIMA",    "Flexible, can evolve",     "High",                "Trend smooth; seasonal has autocorrelation",
  "STL + ARIMA + ETS",    "Flexible, can evolve",     "High",                "Trend autocorrelated; seasonal is level-shifting",
  "ETS",                  "Fixed structure",           "Moderate",            "Short series; no explicit decomposition needed",
  "SARIMA",               "Fixed structure",           "Lower",               "Compact model preferred; longer series"
) |>
  knitr::kable()
```

:::

:::{.fragment}
:::{.callout-important appearance="simple"}
## The guiding principle

No approach dominates universally. Always compare on a held-out test set. The mixed combinations (`stl_ets_arima`, `stl_arima_ets`) are not automatically better — they add flexibility but also complexity. Let the accuracy metrics decide.
:::
:::

# Module 2 Wrap-Up

:::{.content-hidden unless-format="revealjs"}
##
:::

We started Module 2 with a single question: *can we do better than Drift and SNAIVE?*

:::{.incremental}

- We replaced **Drift** with **ETS** — adaptive smoothing that weights recent observations more heavily and handles a wide range of trend and seasonal patterns. ([Exponential Smoothing](../01_ets/ets.qmd))

- We formalized **stationarity**, and learned to use Box-Cox transformations, differencing, and unit root tests to prepare a series for ARIMA modeling. ([Stationarity & Differencing](../02_stationarity/stationarity.qmd))

- We built **ARIMA** models — combining differencing, AR, and MA terms into a unified framework, applied both inside STL and as a standalone seasonal model. ([ARIMA Models](../03_arima/arima.qmd))

- Today, we discovered that `decomposition_model()` lets us **mix ETS and ARIMA freely** across components — and saw that naming specs separately keeps model comparison code clean and reusable across series.

:::

:::{.fragment}
:::{.callout-note appearance="simple"}
## What's still missing?

All models in Module 2 are **univariate** — they only look at the history of the series itself, with no knowledge of the world outside: promotions, holidays, economic conditions, competitor behavior.

In **Module 3**, we give our models that external context by introducing regression with time series errors.
:::
:::
