---
title: "Time Series Decomposition"
format:
  html: default
  revealjs:
    output-file: r_time_series_pres.html
---

:::{.content-visible unless-format="revealjs"}

```{r}
#| label: pkgs
#| message: false

library(tidyverse)
library(fpp3)
library(tidyquant)
library(plotly)
```

:::

# TS Features & Patterns

## 

```{r}
#| label: ts-examples
#| echo: false
#| message: false
#| layout-ncol: 2

p1 <- tq_get("FARTCOIN-USD", from = "2024-01-01") |> 
  as_tsibble(index = date) |>
  autoplot(close, color = "navyblue") +
  labs(
    title = "Fartcoin",
    y = "USD",
    caption = "Source: Yahoo Finance"
  )

ggplotly(p1)  

p2 <- tq_get(
  "MEXLOCOEMORSTM",
  get = "economic.data",
  from = "1985-01-01"
) |>
  mutate(date = yearmonth(date)) |> 
  as_tsibble(index = date) |> 
  autoplot(price, color = "orchid") +
  labs(
    title = "Employment and Unemployment: Original Series for Mexico",
    y = "Percent",
    caption = "Source: Organization for Economic Co-operation and Development via FRED"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) 
 
ggplotly(p2, dynamicTicks = TRUE)

p3 <- us_employment |>
  filter_index("1960 Jan." ~ .) |> 
  filter(Title == "Mining and Logging") |> 
  autoplot(color = "indianred1") +
  labs(
    title = "Mining and Logging Employment in the US",
    y = "Thousands of persons employed",
    caption = "Source = U.S. Bureau of Labor Statistics"
  )

ggplotly(p3, dynamicTicks = TRUE)

p4 <- tq_get(
  "MEXSLRTTO01IXOBM",
  get = "economic.data",
  from = "1985-01-01"
) |>
  mutate(date = yearmonth(date)) |> 
  as_tsibble(index = date) |> 
  autoplot(price, color = "springgreen4") +
  labs(
    title = "Retail Trade, Except of Motor Vehicles and Motorcycles for Mexico",
    y = "Index 2015=100",
    caption = "Organization for Economic Co-operation and Development via FRED"
  ) 
 
ggplotly(p4, dynamicTicks = TRUE)  
```

:::{.notes}
All these time series have different shapes, patterns, and so on. 

:::

## TS Patterns

Time series can have distinct patterns:

- **Trend:** A long-term increase/decrease in the data.

- **Seasonal:** Fluctuations in the time series with a fixed and known period^[A time series can have multiple seasonal patterns.].

- **Cycles:** More commonly known as "Business cycles", refer to rises and falls that are not of a fixed frequency^[They usually last at least 2 years.].



# Transformations & Adjustments



# Time Series Decomposition

## Types of Decompositions

:::{.notes}

A decomposition splits the time series into its underlying components:

- Trend-cycle
- Seasonal pattern(s)

And what's left of it we simply call it a "remainder component".

In general, there are two types of decompositions:
:::

### Additive decomposition

$$
Y_t = T_t + S_t + R_t
$$

### Multiplicative decomposition

$$
Y_t = T_t \times S_t \times R_t \\
$$


# Benchmark Forecasting Methods

# Residual Diagnostics



## Portmanteau Tests of Autocorrelation
