---
title: "Exponential smoothing"
format:
  html:
    other-links:
      - text: FPP3 Chapter 8
        href: https://otexts.com/fpp3/expsmooth.html
  revealjs:
    output-file: ets_pres.html
---


```{r}
#| label: pkgs
#| message: false
#| include: false

library(tidyverse)
library(fpp3)
library(plotly)

source(here::here("R/narsil_theme.R"))
theme_set(theme_narsil())
```

## Introduction

In Module 1, you built your **first complete forecasting model**. That model used:

- **STL decomposition** to separate trend-cycle and seasonal patterns.
- **Drift** to forecast the trend-cycle component (assuming the last observed trend continues linearly).
- **SNAIVE** to forecast the seasonal component (assuming last year's season repeats).

:::{.fragment}

:::{.callout-note appearance="simple"}
## The Module 1 baseline model

```{r}
#| eval: false

tsibble |>
  model(
    baseline = decomposition_model(
      STL(y ~ trend(window = NULL) + season(window = "periodic"), robust = TRUE),
      RW(season_adjust ~ drift()),
      SNAIVE(season_year)
    )
  )
```

This is your **benchmark**. Every model we build from now on must beat it.
:::

:::

::: {.notes}
That's a solid start — but the Drift method is very simple: it just extrapolates the last observed average growth rate forever. **What if the trend is slowing down? Accelerating? Or changing direction?**
:::

:::{.content-hidden unless-format="revealjs"}

##

:::

This module introduces **Exponential Smoothing (ETS)** models — a smarter way to model the trend-cycle component (and eventually, the seasonal component too).

:::{.callout-tip appearance="simple"}
## The upgrade we're making

| Component     | Module 1 (Baseline)  | Module 2 (ETS upgrade)      |
|:--------------|:--------------------:|:---------------------------:|
| Trend-cycle   | Drift (RW + drift)   | **ETS** (adaptive smoothing)|
| Seasonal      | SNAIVE               | SNAIVE (for now)            |
| Decomposition | STL                  | STL (for now)               |

We're making the trend-cycle component **smarter**. Same architecture, better engine.
:::

:::{.content-hidden unless-format="revealjs"}

## {auto-animate=true}
::: {.r-hstack style="justify-content:space-between; align-items:center; margin-top:50px;"}
::: {data-id="mean"  style="background:#2F855A; color:white; padding:10px 14px; border-radius:12px;"}
**Mean**  
$$
\hat y_{T+1\mid T}=\tfrac{1}{T}\sum_{i=1}^T y_i
$$

:::

::: {data-id="axis" style="flex:1; height:10px; margin:0 16px; border-radius:6px; background: linear-gradient(90deg,#2F855A,#2CA1C2,#C0392B);"} 
:::

::: {data-id="naive" style="background:#C0392B; color:white; padding:10px 14px; border-radius:12px;"}
**Naïve**  
$$
\hat y_{T+1\mid T}=y_T
$$

:::
:::


## {auto-animate=true}
::: {.r-hstack style="justify-content:space-between; align-items:center; margin-top:50px;"}
::: {data-id="mean"  style="background:#2F855A; color:white; padding:10px 14px; border-radius:12px; opacity:.9;"}
**Mean** 
:::

::: {data-id="axis" style="flex:1; height:10px; margin:0 16px; border-radius:6px; background: linear-gradient(90deg,#2F855A,#2CA1C2,#C0392B);"} 
:::

::: {data-id="naive" style="background:#C0392B; color:white; padding:10px 14px; border-radius:12px; opacity:.9;"} 
**Naïve** 
:::

::: {data-id="ets" style="position:absolute; left:50%; transform:translateX(-50%); top:20%; background:#2CA1C2; color:white; padding:16px 20px; border-radius:14px; box-shadow:0 6px 18px #0006;"}
**Exponential Smoothing**  
$$
\hat y_{T+1\mid T}=\alpha y_T +  \alpha(1-\alpha)y_{T-1} + \ldots 
$$  

- $\alpha \approx 1$: naïve-like

- $\alpha \approx 0$: mean-like


:::
:::

:::

:::{.notes}

- Exponential smoothing methods are still relatively simple: they're simply weighted averages from historical data. 
  - However, these forecasting methods are widely used in practice, and they can be very effective.

- The exponential smoothing method is a compromise between the mean and naïve methods. It uses all historical data, but it assigns exponentially decreasing weights to older observations.

  - In the *mean* method, all observations are weighted equally (all have the same importance), while in the *naïve* method, only the most recent observation is used for forecasting. (we ignore all previous observations).

- The smoothing parameter $\alpha$ controls the rate of decrease: 
  - when $\alpha$ is close to 1, the method behaves like the naïve method, giving more weight to recent observations; 
  - when $\alpha$ is close to 0, it behaves like the mean method, giving more equal weight to all observations.

:::

:::{.content-hidden unless-format="revealjs"}
##
:::

$$
\hat{y}_{T+1 | T}= \alpha y_{T} + \alpha(1-\alpha) y_{T-1} + \alpha(1-\alpha)^{2} y_{T-2}  + \ldots
$$

where $0\leq \alpha \leq1$ is the smoothing parameter.

  

Table: Weights for different values of $\alpha$ {#tbl-ets_alpha_weights .info .striped .hover .responsive}

|              | $\alpha = 0.2$ | $\alpha = 0.4$ | $\alpha = 0.6$ | $\alpha = 0.8$ |
|--------------|:--------------:|:--------------:|:--------------:|:--------------:|
| $y_t$        | 0.2000         | 0.4000         | 0.6000         | 0.8000         |
| $y_{t-1}$    | 0.1600         | 0.2400         | 0.2400         | 0.1600         |
| $y_{t-2}$    | 0.1280         | 0.1440         | 0.0960         | 0.0320         |
| $y_{t-3}$    | 0.1024         | 0.0864         | 0.0384         | 0.0064         |
| $y_{t-4}$    | 0.0819         | 0.0518         | 0.0154         | 0.0013         |
| $y_{t-5}$    | 0.0655         | 0.0311         | 0.0061         | 0.0003         |



:::{.notes}
- $\alpha$ can be thought of as the *memory* of the time series: The smaller the value of $\alpha$, the longer the memory (i.e., the more past observations are taken into account). 

- Conversely, a larger value of $\alpha$ means a shorter memory, with more emphasis on recent observations. See @tbl-ets_alpha_weights for some examples.
:::

# Exponential smoothing methods

## Simple exponential smoothing (SES)

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = \ell_t \\
\text{Smoothing equation} \quad & \ell_t = \alpha y_t + (1-\alpha)\ell_{t-1}
\end{aligned}
$$

where $\ell_t$ is the level at time $t$.

:::{.callout-note appearance="simple"}
SES has a flat forecast function, so it is appropriate for data with **no trend** or **seasonal** pattern.
:::

::: {.content-hidden unless-format="revealjs"}
## Example: Forecasting Algeria's exports {auto-animate="true"}
:::

::: {.content-visible unless-format="revealjs"}
#### Example: Forecasting Algeria's exports {.unnumbered}
:::

```{r}
#| label: alg-plot
#| code-fold: true

algeria_economy <- global_economy |>
  filter(Country == "Algeria")
  
algeria_economy |> 
  autoplot(Exports)
```

:::{.fragment}

```{r}
#| label: ses

alg_fit <- algeria_economy |>
  model(
    SES = ETS(Exports ~ error("A") + trend("N") + season("N")), #<1>
    Naive = NAIVE(Exports)
  )

alg_fc <- alg_fit |>
  forecast(h = 5)
```

1. We specify `trend("N")` and `season("N")` to indicate that we want a simple exponential smoothing (SES) model, which assumes no trend and no seasonality. The model will estimate the smoothing parameter $\alpha$ automatically.

:::

::: {.content-visible unless-format="revealjs"}

:::{.callout-tip appearance="simple"}
## Benchmark methods for SES

The **mean** and **naïve** methods are typically the best fit as benchmark methods when using SES.

:::

:::

::: {.fragment}

::: {.callout-note collapse="true"}
## Obtaining the `report()` of a model 
```{r}
#| label: ses-report

alg_fit |> 
  select(SES) |> 
  report()        #<1>
```

1. The `report()` function allows us to see a model's report (the time series modeled, the model used, the estimated parameters, and more). It needs a $1 \times 1$ dimension `mable`^[(i.e., a `mable` containing only one model and one time series.)]. 

:::

:::

::: {.content-hidden unless-format="revealjs"}
## The fitted values & forecast
:::

```{r}
#| label: ses-plot
#| echo: false

alg_fc |> 
  filter(.model == "SES") |> 
  autoplot(algeria_economy) +
  geom_line(aes(y = .fitted), color = "darkorange",
            data = alg_fit |> select(SES) |> augment()) +
  labs(
    x = "",
    y = "% of GDP", 
    title = "Forecasting Algeria's exports using SES"
  )
```

::: {.content-hidden unless-format="revealjs"}
##
:::

Comparing the SES and Naive forecasts:

```{r}
#| label: ses-v-naive-plot
#| echo: false

alg_fc |> 
  autoplot(algeria_economy |> filter_index(2005 ~ .), level = NULL) +
  labs(
    x = "",
    y = "% of GDP",
    title = "Forecasting using SES vs. Naive",
    caption = "Why is the SES forecast below the Naive here?"
  )
```


::: {.content-hidden unless-format="revealjs"}
# Methods with trend
:::

::: {.content-visible unless-format="revealjs"}
## Methods with trend
:::

::: {.content-hidden unless-format="revealjs"}
## Holt's linear trend
:::

::: {.content-visible unless-format="revealjs"}
### Holt's linear trend
:::

:::{.notes}
We can extend SES models to allow our forecasts to include trend in the data. We need to add a new smoothing parameter $\beta^*$, and its corresponding smoothing equation:
:::

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = \ell_t + hb_t \\
\text{Level equation} \quad & \ell_t = \alpha y_t + (1-\alpha)\ell_{t-1}\\
\text{Trend equation} \quad & b_t = \beta^*(\ell_t - \ell_{t-1}) + (1-\beta^*)b_{t-1}
\end{aligned}
$$

where $b_t$ is the growth (or slope) at time $t$.

:::{.fragment}

:::{.callout-tip collapse="true"}
## When to use Holt's linear trend method
- Holt's linear trend method is appropriate for data with a **linear trend** but **no seasonal** pattern.
- The proper benchmark method to compare against is the **drift** method. 
:::

:::

:::{.notes}
Let's see an example using Holt's linear trend method to forecast Brazil's population.
:::

::: {.content-visible unless-format="revealjs"}
#### Example: Forecasting Brazil's population {.unnumbered}

```{r}
#| label: bra-pop

bra_economy <- global_economy |> 
  filter(Code == "BRA") |> 
  mutate(Pop = Population / 1e6)

bra_economy |> 
  autoplot(Pop)
```

:::

::: {.content-hidden unless-format="revealjs"}
## Example: Forecasting Brazil's population {auto-animate="true"}

```{r}
#| label: bra-pop-plot
#| echo: false

bra_economy <- global_economy |> 
  filter(Code == "BRA") |> 
  mutate(Pop = Population / 1e6)

bra_economy |> 
  autoplot(Pop)
```

:::


::: {.content-hidden unless-format="revealjs"}

## Example: Forecasting Brazil's population {auto-animate="true"}

```{r}
#| label: holt_1

bra_fit <- bra_economy |> 
  model(
    Holt  = ETS(Pop ~ error("A") + trend("A") + season("N"))
  )

bra_fit
```

## Example: Forecasting Brazil's population {auto-animate="true"}

```{r}
#| label: holt_2

bra_fit <- bra_economy |> 
  model(
    Holt  = ETS(Pop ~ error("A") + trend("A") + season("N")),
    Drift = RW(Pop ~ drift())
  )

bra_fit
```

## Example: Forecasting Brazil's population {auto-animate="true"}

```{r}
#| label: holt_3

bra_fit <- bra_economy |> 
  model(
    Holt  = ETS(Pop ~ error("A") + trend("A") + season("N")),
    Drift = RW(Pop ~ drift())
  )

bra_fit |>  
  select(Holt) |>  
  report()
```


:::

::: {.content-hidden unless-format="revealjs"}
## Example: Forecasting Brazil's population {auto-animate="true"}
:::


```{r}
#| label: holt_full

bra_fit <- bra_economy |> 
  model(
    Holt  = ETS(Pop ~ error("A") + trend("A") + season("N")), #<1>
    Drift = RW(Pop ~ drift())
  )

bra_fit |>  
  select(Holt) |>  
  report()

bra_fc <- bra_fit |>  
  forecast(h = 15)
```

1. We specify `trend("A")` to indicate that we want a linear trend. The model will estimate the smoothing parameters $\alpha$ and $\beta^*$ automatically.

```{r}
#| label: holt_full_plt
#| code-fold: true
bra_fc |>
  autoplot(bra_economy, level = NULL) +
  labs(title = "Brazilian population", y = "Millions") +
  guides(colour = guide_legend(title = "Forecast"))
```

::: {.content-hidden unless-format="revealjs"}
## Damped trend
:::

::: {.content-visible unless-format="revealjs"}
### Damped trend
:::

:::{.notes}
- Holt's linear trend method assume that the trend will continue indefinitely at the same rate. However, in many real-world scenarios, this assumption may not hold true. This methods tend to overestimate (*or underestimate*) long-term forecasts when the trend is strong.

- We can include a damping parameter $\phi$, which reduces the trend over time.
:::

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = \ell_t +  (\phi + \phi^2 + \ldots + \phi^h) b_t \\
\text{Level equation} \quad & \ell_t = \alpha y_t + (1 - \alpha) (\ell_{t-1} + \phi b_{t-1}) \\
\text{Trend equation} \quad & b_t = \beta^*(\ell_t-\ell_{t-1}) + (1-\beta^*)\phi b_{t-1}
\end{aligned}
$$

where $0 < \phi < 1$^[In practice, we restrict $0.8 \leq \phi \leq 0.98$ because the damping effect would be too great for smaller values than 0.8 and almost non distinguishable from a linear trend for greater values than 0.98.] is the damping parameter.

:::{.fragment}

::: {.callout-caution collapse="true"}
## What would happen if $\phi = 1$? What about if $\phi = 0$?
- If $\phi = 1$, the model reduces to Holt's linear trend method, meaning the trend continues indefinitely at the same rate.
- If $\phi = 0$, the trend component is completely eliminated, and the model behaves like simple exponential smoothing (SES), where forecasts are based solely on the level component without any trend influence.
:::

:::

::: {.content-hidden unless-format="revealjs"}

## Example: Forecasting Brazil's population (continued) {auto-animate="true"}

```{r}
#| label: damped_1

bra_economy |> 
  model(
    Holt   = ETS(Pop ~ error("A") + trend("A") + season("N"))
  )
```

## Example: Forecasting Brazil's population (continued) {auto-animate="true"}

```{r}
#| label: damped_2

bra_economy |> 
  model(
    Holt   = ETS(Pop ~ error("A") + trend("A") + season("N")),
    Damped = ETS(Pop ~ error("A") + trend("Ad", phi = 0.9) + season("N"))
  )
```

:::

::: {.content-hidden unless-format="revealjs"}
## Example: Forecasting Brazil's population (continued) {auto-animate="true"}
:::

::: {.content-visible unless-format="revealjs"}
#### Example: Forecasting Brazil's population (continued) {.unnumbered}
:::

```{r}
#| label: damped_full

bra_economy |> 
  model(
    Holt   = ETS(Pop ~ error("A") + trend("A") + season("N")),
    Damped = ETS(Pop ~ error("A") + trend("Ad", phi = 0.9) + season("N")) #<1>
  ) |> 
  forecast(h = 15) |> 
  autoplot(bra_economy, level = NULL) +
  labs(title = "Brazilian population",
       y = "Millions") +
  guides(colour = guide_legend(title = "Forecast"))
```

1. We specify `trend("Ad")` to indicate that we want a damped trend, and `phi = 0.9` sets the damping parameter to 0.9. We could also let the model estimate $\phi$ automatically by omitting the `phi` argument.

:::{.content-hidden unless-format="revealjs"}

# Handling seasonality: two strategies
:::

:::{.content-visible unless-format="revealjs"}
## Handling seasonality: two strategies
:::

:::{.notes}
So far, we've improved the trend-cycle component of our baseline model. But we haven't touched the seasonal component yet — SNAIVE is still handling it. There are two ways to bring ETS into seasonal data, and they lead to different model families.
:::

::: {.content-hidden unless-format="revealjs"}
##
:::



:::{.callout-tip appearance="simple"}
## Two ways to handle seasonality with ETS

| Strategy | How it works | When to prefer it |
|:---------|:-------------|:------------------|
| **Decompose first** | Use STL to extract seasonality, then fit ETS on the seasonally adjusted series | Seasonal pattern is complex, changes over time, or you want explicit control |
| **Model directly** | Use a full ETS model with a seasonal component (Holt-Winters) | Seasonal pattern is stable; simpler, self-contained model |

:::

:::{.content-hidden unless-format="revealjs"}
## Strategy 1: STL + ETS via `decomposition_model()`
:::

::: {.content-visible unless-format="revealjs"}
### Strategy 1: STL + ETS via `decomposition_model()`
:::

:::{.notes}

In Module 1 you used `decomposition_model()` to pair STL with Drift + SNAIVE. The same wrapper accepts **any** model for the seasonally adjusted series — including ETS.

:::

```{r}
#| label: aus-holidays-data
#| code-fold: true

aus_holidays <- tourism |> 
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips))

aus_holidays |> autoplot(Trips) +
  labs(title = "Australian holiday trips",
       y = "Overnight trips (millions)")
```

::: {.content-hidden unless-format="revealjs"}

## Strategy 1: STL + ETS {auto-animate="true"}

```{r}
#| label: stl_ets_1
#| code-annotation: below

stl_ets_add_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  SNAIVE(season_year), # <1>
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))
)
```

1. If you want to model the series' seasonality using `SNAIVE`, you can simply **not specify it** and R will take it as default.

## Strategy 1: STL + ETS {auto-animate="true"}

```{r}
#| label: stl_ets_2

stl_ets_add_spec <- decomposition_model(          
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N")) 
)
```


## Strategy 1: STL + ETS {auto-animate="true"}

```{r}
#| label: stl_ets_3

stl_ets_add_spec <- decomposition_model(           
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))
)

stl_ets_mul_spec <- decomposition_model(     
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("M") + trend("A") + season("N"))
)
```

:::

::: {.content-hidden unless-format="revealjs"}
## Strategy 1: STL + ETS {auto-animate="true"}
:::

```{r}
#| label: stl_ets_full

stl_ets_add_spec <- decomposition_model(                                        #<1>
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))                    #<2>
)

stl_ets_mul_spec <- decomposition_model(                                        #<3>
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("M") + trend("A") + season("N"))                    #<4>
)

aus_holidays |> 
  model(
    stl_ets_add = stl_ets_add_spec,
    stl_ets_mul = stl_ets_mul_spec
  ) |> 
  forecast(h = "2 years") |> 
  autoplot(aus_holidays, level = NULL) +
  labs(title = "STL + ETS forecasts",
       y = "Overnight trips (millions)") +
  scale_color_brewer(type = "qual", palette = "Dark2")
```

1. Save the decomposition spec as a standalone object — this keeps `model()` clean and allows reuse.
2. The ETS model is fitted on `season_adjust` (the STL-extracted seasonally adjusted series). We set `season("N")` because STL has already removed the seasonal component; SNAIVE handles it implicitly via `decomposition_model()`.
3. A multiplicative error version of the same spec.
4. Multiplicative errors (`error("M")`) are often better when the variance of the series grows with its level.

:::{.callout-note appearance="simple"}
## What `decomposition_model()` does with the seasonal component

When you pass only one sub-model (the ETS on `season_adjust`), `decomposition_model()` automatically uses **SNAIVE** for the seasonal component. You can override this by passing a second model explicitly, as in the Module 1 baseline.
:::

:::{.content-hidden unless-format="revealjs"}
## Strategy 2: Holt-Winters
:::

::: {.content-visible unless-format="revealjs"}
### Strategy 2: Holt-Winters
:::

Instead of decomposing first, Holt-Winters extends ETS to model level, trend, and seasonality **simultaneously**.

::: {.content-hidden unless-format="revealjs"}
## HW — Additive
:::

::: {.content-visible unless-format="revealjs"}
#### HW — Additive
:::

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = \ell_t + hb_t + s_{t+h-m(k+1)} \\
\text{Level equation} \quad & \ell_t = \alpha (y_t - s_{t-m}) + (1 - \alpha) (\ell_{t-1} + b_{t-1}) \\
\text{Trend equation} \quad & b_t = \beta^*(\ell_t-\ell_{t-1}) + (1-\beta^*) b_{t-1} \\
\text{Seasonal equation} \quad & s_t = \gamma(y_t - \ell_{t-1} - b_{t-1}) + (1-\gamma)s_{t-m}
\end{aligned}
$$

where $s_t$ is the seasonal component at time $t$, $m$ is the period of the seasonality^[e.g., $m=4$ for quarterly data, $m=12$ for monthly data, ...], and $k = \lfloor (h-1)/m \rfloor$.

::: {.content-hidden unless-format="revealjs"}
## HW — Multiplicative
:::

::: {.content-visible unless-format="revealjs"}
#### HW — Multiplicative
:::

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = (\ell_t + hb_t) s_{t+h-m(k+1)} \\
\text{Level equation} \quad & \ell_t = \alpha \frac{y_t}{s_{t-m}} + (1 - \alpha)(\ell_{t-1} + b_{t-1}) \\
\text{Trend equation} \quad & b_t = \beta^*(\ell_t-\ell_{t-1}) + (1-\beta^*) b_{t-1} \\
\text{Seasonal equation} \quad & s_t = \gamma \frac{y_t}{\ell_{t-1} + b_{t-1}} + (1-\gamma)s_{t-m}
\end{aligned}
$$

:::{.callout-tip collapse="true"}
## Additive vs. multiplicative: how to choose
- Use **additive** when seasonal fluctuations are roughly constant in size over time.
- Use **multiplicative** when seasonal variation grows or shrinks with the level of the series.
- When in doubt, look at the time plot: if the seasonal swings widen as the series rises, go multiplicative.
:::

::: {.content-hidden unless-format="revealjs"}

## Example: Australian holiday trips {auto-animate="true"}

```{r}
#| label: hw_1

aus_fit_hw <- aus_holidays |> 
  model(
    hw_add = ETS(Trips ~ error("A") + trend("A") + season("A"))
  )
```

## Example: Australian holiday trips {auto-animate="true"}

```{r}
#| label: hw_2

aus_fit_hw <- aus_holidays |> 
  model(
    hw_add = ETS(Trips ~ error("A") + trend("A") + season("A")),
    hw_mul = ETS(Trips ~ error("M") + trend("A") + season("M"))
  )
```

## Example: Australian holiday trips {auto-animate="true"}

:::

::: {.content-visible unless-format="revealjs"}
#### Example: Forecasting Australian holiday trips {.unnumbered}
:::

```{r}
#| label: hw_full

aus_fit_hw <- aus_holidays |> 
  model(
    hw_add = ETS(Trips ~ error("A") + trend("A") + season("A")), #<1>
    hw_mul = ETS(Trips ~ error("M") + trend("A") + season("M"))  #<2>
  )
```

1. `error("A")` and `season("A")`: additive errors and additive seasonality. Seasonal fluctuations are treated as constant in size.
2. `error("M")` and `season("M")`: multiplicative errors and multiplicative seasonality. Seasonal fluctuations scale with the level of the series.

```{r}
#| label: hw_full_plt
#| code-fold: true
aus_fit_hw |>
  forecast(h = "3 years") |>
  autoplot(aus_holidays, level = NULL) +
  labs(
    title = "Holt-Winters forecasts: additive vs. multiplicative",
    y = "Overnight trips (millions)",
    caption = "Can you spot any differences between both forecasts?"
  ) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  guides(colour = guide_legend(title = "Forecast"))
```

::: {.callout-note collapse="true"}
## The `tidy()` function for models

```{r}
#| label: tidy

aus_fit_hw |> 
  tidy()  #<1>
```

1. The `tidy()` function allows us to see the estimated parameters of each model in a [tidy](https://r4ds.hadley.nz/data-tidy.html) table.
:::

::: {.content-hidden unless-format="revealjs"}
## Holt-Winters' damped method
:::

::: {.content-visible unless-format="revealjs"}
#### Holt-Winters' damped method
:::

:::{.notes}

- Similar to Holt's linear trend method, we can also include a damping parameter $\phi$ in the Holt-Winters method to reduce the trend over time.
  - This can be done on both additive or multiplicative seasonal models.
- A configuration that has proven to be robust and accurate in practice is to use a multiplicative seasonal component with a damped trend.

:::

$$
\begin{aligned}
\text{Forecast equation} \quad & \hat{y}_{t+h|t} = [\ell_t +(\phi + \phi^2 + \ldots + \phi^h)b_t] s_{t+h-m(k+1)} \\
\text{Level equation} \quad & \ell_t = \alpha \frac{y_t}{s_{t-m}} + (1 - \alpha)(\ell_{t-1} + b_{t-1}) \\
\text{Trend equation} \quad & b_t = \beta^*(\ell_t-\ell_{t-1}) + (1-\beta^*) \phi b_{t-1} \\
\text{Seasonal equation} \quad & s_t = \gamma \frac{y_t}{\ell_{t-1} + \phi b_{t-1}} + (1-\gamma)s_{t-m}
\end{aligned}
$$

::: {.content-hidden unless-format="revealjs"}

## Example: Forecasting daily pedestrian traffic {auto-animate="true"}

:::

::: {.content-visible unless-format="revealjs"}

#### Example: Forecasting daily pedestrian traffic {.unnumbered}

:::

```{r}
#| label: hw_damped_setup
#| code-fold: true

sth_cross_ped <- pedestrian |>
  filter(Date >= "2016-07-01", Sensor == "Southern Cross Station") |>
  index_by(Date) |>
  summarise(Count = sum(Count) / 1000)
```

```{r}
#| label: hw_damped
#| code-line-numbers: "|4"

sth_cross_ped |>
  filter(Date <= "2016-07-31") |>
  model(
    hw = ETS(Count ~ error("M") + trend("Ad") + season("M"))
  ) |>
  forecast(h = "2 weeks") |>
  autoplot(sth_cross_ped |> filter(Date <= "2016-08-14")) +
  labs(title = "Daily traffic: Southern Cross",
       y="Pedestrians ('000)")
```

:::{.callout-tip appearance="simple"}

The setup `ETS(y ~ error("M") + trend("Ad") + season("M"))` is often a robust choice for seasonal data with trend.

:::

::: {.content-hidden unless-format="revealjs"}

# Automatic ETS selection

:::

::: {.content-visible unless-format="revealjs"}

## Automatic ETS selection

:::

:::{.notes}

- Using `fable`, we can automatically select the best ETS model for our data using the `ETS()` function. We achieve this by **not specifying** the error, trend, or seasonal components.
- You can also get a *semi-automatic* selection by specifying some components and leaving others to be selected automatically, or by providing a set of possible values for a component.
- Automatic selection is especially useful when we have multiple time series to model, as it allows us to fit different ETS models to each series.
:::

::: {.content-hidden unless-format="revealjs"}

## Example: Forecasting daily pedestrian traffic (continued) {auto-animate="true"}

```{r}
#| label: auto_ets_js
#| code-line-numbers: "10-11"
#| paged.print: false

ped_fit <- pedestrian |>
  filter(
    Date >= "2016-07-01",
    Sensor != "Birrarung Marr"
  ) |>
  index_by(Date) |>
  group_by_key() |>
  summarise(Count = sum(Count) / 1000) |>
  model(
    ETS_auto = ETS(Count),
    ets_with_trend = ETS(Count ~ trend(c("A", "Ad"))),
  )

ped_fit
```


:::

::: {.content-visible unless-format="revealjs"}

#### Example: Forecasting daily pedestrian traffic (continued) {.unnumbered}

```{r}
#| label: auto_ets
#| paged.print: false

ped_fit <- pedestrian |> 
  filter(
    Date >= "2016-07-01",
    Sensor != "Birrarung Marr"
  ) |> 
  index_by(Date) |>
  group_by_key() |> 
  summarise(Count = sum(Count)/1000) |> 
  model(
    ETS_auto       = ETS(Count),                       #<1>
    ets_with_trend = ETS(Count ~ trend(c("A", "Ad"))), #<2>
  )

ped_fit                                                #<3>
```
1. Leaving the right-hand side of the formula empty allows `fable` to automatically select the best ETS model for each time series.
2. Here, we specify that the trend component can be either additive or damped, while the error and seasonal components will be selected automatically.
3. The output shows the selected model for each sensor.

:::


# Model comparison

:::{.notes}
Now we put all five models side by side on a proper train/test split. The goal is not just to see which number wins, but to understand *why* some approaches work better for this particular series.
:::

::: {.content-hidden unless-format="revealjs"}

## Train/test split

:::

::: {.content-visible unless-format="revealjs"}

### Train/test split

:::

We now evaluate all five models on the same holdout period: the **last 8 quarters** of `aus_holidays` (2 years).

```{r}
#| label: train-test-split

h <- 8

holidays_train <- aus_holidays |> 
  slice_head(n = nrow(aus_holidays) - h)   #<1>

holidays_test <- aus_holidays |> 
  slice_tail(n = h)                         #<2>
```

1. All observations except the last 8 quarters go into training.
2. The last 8 quarters are held out for evaluation.

```{r}
#| label: train-test-split-plt
#| code-fold: true
holidays_train |>
  autoplot(Trips) +
  autolayer(holidays_test, Trips, colour = "#C0392B", linetype = "dashed") +
  labs(
    title = "Australian holiday trips: train/test split",
    subtitle = glue::glue(
      "Training: {min(holidays_train$Quarter)} – {max(holidays_train$Quarter)}  |  Test: {min(holidays_test$Quarter)} – {max(holidays_test$Quarter)}"
    ),
    y = "Overnight trips (millions)"
  )
```

::: {.content-hidden unless-format="revealjs"}

## Fitting all five models {auto-animate="true"}

```{r}
#| label: all-models-fit-1
#| eval: false

# Module 1 baseline spec (reused from Module 1)
baseline_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  RW(season_adjust ~ drift())
)
```

## Fitting all five models {auto-animate="true"}

```{r}
#| label: all-models-fit-2
#| eval: false
#| code-line-numbers: "7-11"

# Module 1 baseline spec (reused from Module 1)
baseline_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  RW(season_adjust ~ drift())
)

# STL + ETS specs (defined earlier in this document)
stl_ets_add_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))
)
```

## Fitting all five models {auto-animate="true"}

```{r}
#| label: all-models-fit-3
#| eval: false
#| code-line-numbers: "13-16"

# Module 1 baseline spec (reused from Module 1)
baseline_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  RW(season_adjust ~ drift())
)

# STL + ETS specs (defined earlier in this document)
stl_ets_add_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))
)

stl_ets_mul_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("M") + trend("A") + season("N"))
)
```

## Fitting all five models {auto-animate="true"}

:::

::: {.content-visible unless-format="revealjs"}

### Fitting all five models

:::

```{r}
#| label: all-models-fit
#| paged.print: false

# Module 1 baseline spec (reused from Module 1)
baseline_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  RW(season_adjust ~ drift())
)

# STL + ETS specs (defined earlier in this document)
stl_ets_add_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("A") + trend("A") + season("N"))
)

stl_ets_mul_spec <- decomposition_model(
  STL(Trips ~ trend(window = 7) + season(window = "periodic"), robust = TRUE),
  ETS(season_adjust ~ error("M") + trend("A") + season("N"))
)

holidays_fit <- holidays_train |> 
  model(
    baseline    = baseline_spec,                                             #<1>
    stl_ets_mul = stl_ets_mul_spec,                                          #<3>
    stl_ets_add = stl_ets_add_spec,                                          #<2>
    hw_add      = ETS(Trips ~ error("A") + trend("A") + season("A")),        #<4>
    hw_mul      = ETS(Trips ~ error("M") + trend("A") + season("M"))         #<5>
  )

holidays_fit
```

1. Module 1 baseline: STL + Drift + SNAIVE. The model every new approach must beat.
2. STL decomposition, then additive ETS on the seasonally adjusted series.
3. STL decomposition, then multiplicative ETS on the seasonally adjusted series.
4. Holt-Winters additive: models level, trend, and seasonality jointly with additive components.
5. Holt-Winters multiplicative: seasonal fluctuations scale with the level of the series.

::: {.content-hidden unless-format="revealjs"}

## Forecast plot

:::

::: {.content-visible unless-format="revealjs"}

### Forecast plot

:::

```{r}
#| label: all-models-fc
#| code-fold: true

holidays_fc <- holidays_fit |> 
  forecast(h = h)

holidays_fc |> 
  autoplot(
    aus_holidays |> filter_index("2012 Q1" ~ .),
    level = NULL
  ) +
  labs(
    title = "ETS model comparison: 5 models, 8-quarter horizon",
    y = "Overnight trips (millions)"
  ) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  guides(colour = guide_legend(title = "Model"))
```

::: {.content-hidden unless-format="revealjs"}

## Accuracy

:::

::: {.content-visible unless-format="revealjs"}

### Accuracy

:::

:::{.content-visible unless-format="revealjs"}

:::{.callout-warning appearance="simple"}
## Pass the full dataset to `accuracy()`

`accuracy()` must receive the **full original tsibble** (not just the test set) as its second argument. MASE and RMSSE denominators are computed from naive errors on the **training data** — passing only the test set makes those metrics undefined.
:::

:::


```{r}
#| label: accuracy
#| df-print: tibble

holidays_accu <- holidays_fc |> 
  accuracy(aus_holidays)           #<1>

holidays_accu |> 
  select(.model, RMSE, MAE, MAPE, RMSSE, MASE) |> 
  arrange(RMSSE)
```

1. Always pass the **full** tsibble here, not `holidays_test`. `fable` handles the train/test split internally using the model's training window.

::: {.content-hidden unless-format="revealjs"}

## What does the comparison show?

:::

::: {.content-visible unless-format="revealjs"}

### What does the comparison show?

:::

A few questions worth discussing after looking at the accuracy table:

- Do the ETS models beat the Module 1 baseline? If not, what does that tell us?
- Does the choice between additive and multiplicative matter more for STL + ETS or for Holt-Winters?
- Look at the time plot again: does the seasonal variation grow with the level? Does that match what the accuracy numbers suggest?

:::{.fragment}

:::{.callout-tip collapse="true"}
## Interpreting the results

If `hw_mul` or `stl_ets_mul` win, it is consistent with the visual evidence: the seasonal swings in `aus_holidays` do appear to widen slightly over time, which multiplicative models handle better.

If the baseline is competitive, it is a reminder that more parameters do not automatically mean better forecasts — and that SNAIVE for the seasonal component is often harder to beat than it looks.
:::

:::

# In summary

## The lineup of exponential smoothing methods

Table: ETS component combinations (trend × seasonal) {#tbl-ets-grid .primary .striped .hover .responsive}

| Trend component              | N (None)   | A (Additive)    | M (Multiplicative) |
|:-----------------------------|:----------:|:---------------:|:------------------:|
| $N$ **(None)**               | $(N,N)$,   | $(N,A)$         | $(N,M)$            |
| $A$ **(Additive)**           | $(A,N)$,   | $(A,A)$         | $(A,M)$            |
| $A_d$ **(Additive damped)**  | $(A_d,N)$, | $(A_d, A)$      | $(A_d,M)$          |



:::{.fragment}

Table: Names of some popular ETS models {#tbl-ets-names .secondary .striped .hover .responsive}

| Notation | Method                             |
|:--------:|:----------------------------------:|
|$(N,N)$   | Simple Exponential Smoothing (SES) |
|$(A,N)$   | Holt's Linear Trend                |
|$(A_d,N)$ | Additive damped Trend              |
|$(A,A)$   | Holt-Winters' Additive             |
|$(A,M)$   | Holt-Winters' Multiplicative       |
|$(A_d,M)$ | Holt-Winters' damped               |

:::

:::{.content-hidden unless-format="revealjs"}
##
:::

- Exponential smoothing methods are a family of forecasting methods that use **weighted averages** of past observations to make forecasts.
- The **weights decrease exponentially** for older observations, controlled by smoothing parameters.
- Different configurations of ETS models can be used to handle various data patterns, including trend and seasonality.
- The choice of model components (`error(c("A", "M"))`, `trend(c("N", "A", "Ad"))`, `season(c("N", "A", "M"))`) should be based on the characteristics of the data — start by looking at the time plot.
- When data has both trend and seasonality, two strategies are available: **decompose first** (STL + ETS) or **model jointly** (Holt-Winters). Neither dominates universally.
- Automatic ETS selection can be a powerful tool for fitting models to multiple time series efficiently.
